### 2015-07-31 paper.tex
Ugh, fuck markdown let's do this in latex.

\documentclass[twocolumn]{article}
\usepackage[T1]{fontenc}
\usepackage[cm-default]{fontspec}
\usepackage{lmodern}

\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{sectsty}
\usepackage{ctable}
\usepackage{hyperref}

%\setromanfont{Adagio_Serif Light}
%\setsansfont[BoldFont={Adagio_Sans}]{Adagio_Sans Light}
%\allsectionsfont{\normalfont\sffamily\bfseries}

\title{Isolator: accurate and consistent analysis of RNA-Seq experiments}
\author{Daniel C. Jones \and Kavitha T. Kuppusamy \and Xinxia Peng \and Michael G. Katze \and Hannele Ruohola-Baker \and Walter L. Ruzzo}
\date{}

\begin{document}
\maketitle

\begin{abstract}
    While RNA-Seq has enabled great progress towards the goal of wide-scale
    isoform-level mRNA quantification, short reads have limitations when
    resolving complex and similar sets of isoforms. As a result, estimates
    of isoform abundance, compared to those at the gene level, carry far
    more uncertainty. When confronted with this uncertainty, commonly used
    methods produce estimates that are unstable; small perturbations in the
    data often produce dramatically different results. We introduce a new
    method, Isolator, which analyzes all samples in an experiment in unison
    using a simple Bayesian hierarchical model. Combined with aggressive bias
    correction, it produces estimates that are simultaneously accurate and
    stable. In a comprehensive comparison of accuracy and stability, we show
    that this property is unique to Isolator. We further demonstrate that
    the approach of modeling an entire experiment enables new analyses by
    examining splicing monotonicity across several time points in the
    development of human cardiomyocyte cells.
\end{abstract}

\section{Introduction}\label{introduction}

Since the introduction of RNA-Seq it has rapidly become a preferred
method of studying gene expression, having proved to be reliable,
reproducible, and increasingly affordable. In principle, RNA-Seq enables
measurements of expression at a higher resolution than has previously
been possible or practical at the same scale. Rather than an often
vaguely defined notion of a gene, expression of specific isoform or
exons can be measured. In practice this promise is difficult to realize.
There are no standardized names or symbols for isoforms, annotations are
often incomplete or in disagreement, and methods designed to detect
changes in splicing rarely concur (TODO: cite).

Because a short read is often compatible with many isoforms, the signal
produced from sequencing must be deconvoluted, requiring a certain level
of algorithmic sophistication. This problem was tackled early on and
with great success by Cufflinks, and indeed many of the methods
developed in the interim have followed from the same basic model of
RNA-Seq in which transcripts are represented as particular distributions
over possible reads, and the problem is then to infer mixing
coefficients (relative expression values) inducing a mixture model that
best explains the data.

We built on this work, implementing a new method called Isolator. Though
it retains the same underlying model of RNA-Seq, we approach the problem
from the perspective that isoform-level estimates possess a degree
uncertainty that renders maximum likelihood point-estimates alone a
fundamentally inadequate solution to the problem. As the purpose of gene
expression studies is to compare expression between multiple samples,
our goal, beyond maximizing a narrow definition of accuracy, is to
provide a tool that effectively accounts for this uncertainty in a
coherent and reliable manner.

Rather than estimating expression of individual samples, Isolator uses a
hierarchical model of an entire experiment. In addition to inferring
expression values for the individual samples, we introduce parameters
representing condition-wise and experiment-wise expression and splicing,
as well as per-transcript variance parameters that are shared across
conditions. As expression studies often use only a small number of
replicates, shared variance parameters allows us to make more accurate
estimates of biological variability.

An efficient MCMC algorithm is used to generate samples over the
parameters of this model, which are saved in an HDF5 based format. The
output from the sampler can then be processed to produce point
estimates, credible intervals, posterior probabilities, and diagnostics.
Though designed to be run on multiple samples concurrently, Isolator can
be run as a conventional tool on individual samples.

The broad effect of this model is to encode the common assumption that
genes tend to be similarly expressed and similarly spliced between
replicates, and to a lesser extent between conditions. In the absence of
sufficient evidence to the contrary, this informative prior shrinks
estimates towards a baseline of no change, producing more conservative
estimates of effect size than one would obtain by using, for example, a
flat Dirichlet prior. This feature of the model is particularly
important for robust treatment of low-expression genes. With few reads
there is little power to determine splicing patterns. Considering the
samples in isolation will often produce very different estimates.
Statistical tests based on these estimates, without considering the full
posterior distribution, make approximations by either either
disregarding the relative or compositional nature of the data and
assuming independence, or by relying on asymptotic properties that may
not hold when a gene is not deeply sequenced.

In addition to approach to we attempt to aggressively correct for
technical effects. The naive model of RNA-Seq assumes fragments are
sampled uniformly at random from transcripts in proportion to their
abundance. Several have observed that this uniformity assumption is far
from an accurate description of real sequencing data (TODO: cite).
Though severe at the nucleotide resolution, gene level expression
estimates are only moderately affected by this bias, as it tends to
average out over long transcripts. When one considers expression at the
isoform level, attempting to correct for this nonuniformity becomes
vital, as the difference between two isoforms is often only a few
nucleotides.

We previously developed an efficient and versatile model to account for
sequence specific bias that commonly occur at the ends of fragments
(TODO: cite). In Isolator, we supplemented this with methods accounting
for 3' bias, fragment GC-content bias, and positional bias caused by
end-effects of fragmentation.

Isolator is available under a permissive open source license at:
https://github.com/dcjones/isolator

\section{Results}\label{results}

The convention for evaluating the accuracy of RNA-Seq analysis methods
has consisted chiefly of comparing point estimate accuracy in
simulations, supplemented with comparisons to qPCR or other
technologies. Tuning methods exclusively to maximize accuracy on
simulated data creates the risk of solving only an idealized
mathematical model of RNA-Seq while disregarding noise, bias, and
nonuniformity. As no gold standard for whole-transcriptome isoform-level
estimation exists, we adopt multi-faceted approach, with a focus
measuring the stability or consistency of estimates, along with the
accuracy.

Comparisons of gene or transcript expression is sensitive to how one
chooses to measure similarity or distance. Because gene expression
typically varies across several orders of magnitude, measurements like
the Pearson correlation, $L_1$ or $L_2$ distance, or root-mean-square
error are often dominated by a small subset of the most highly expressed
genes. Recognizing this, a popular alternative is the Spearman rank
correlation, which conversely can overemphasize the majority of
transcripts which have no, or very low expression. These issues are
discussed in depth by Lovell et al (Lovell et al. 2015), who propose
measurements of proportionality as a more meaningful metric. We adopt
their ``proportionality correlation'', defined as
$\text{pcor}(x, y) = 2\text{cov}(\log(x), \log(y)) / (\text{var}(\log(x)) + \text{var}(\log(y)))$.
Similar to other measures of correlation, proportionality correlation
varies from -1 to 1, with 1 indicating perfectly proportionality, and -1
perfect reciprocality. Zeros are accounted for by adding 0.1 to all TPM
values.

Many methods have been proposed to aid in the analysis of RNA-Seq
experiments. We limit our discussion specifically to those that seek to
estimate the relative abundance a set of known transcripts. These
include Cufflinks, RSEM, eXpress, BitSeq, Sailfish, Salmon, BitSeq, and
Kallisto. RSEM was used to produce both maximum likelihood and posterior
mean estimates, which we label ``RSEM/ML'' and ``RSEM/PM'' respectively.
This is not an exhaustive account of such methods, but does represent a
wide variety of popular approaches.

\subsection{Agreement with qPCR and spike-in
controls}\label{agreement-with-qpcr-and-spike-in-controls}

Spike-in controls and qPCR are far more limited than RNA-Seq, measuring
only the abundance of specific loci in the case of qPCR, or known
proportions of simple artificial transcripts with spike-in controls.
However, both are considered reliable enough to serve as proxy gold
standards for gene-level expression estimates.

We use data generated by the SEQC project, consisting of four reference
samples (labeled A, B, C, D). The RNA-seq methods were run using
approximately 25 million reads per sample, obtaining estimates that were
then compared to qPCR in Tables TODO and TODO, and the known ERCC
spike-in mixtures in Table TODO. In each of these we find that Isolator
produces the highest correlation of the methods compared, though
sometimes by negligible margin. Despite two different technologies, and
two different sets of genes, the qPCR benchmarks agree closely, ranking
the methods nearly identically. Cufflinks, eXpress, Salmon, and RSEM/ML
typically perform similarly while BitSeq, Kallisto, Sailfish, and
RSEM/PM trail slightly. Results are similar with ERCC spike-in controls,
the notable exceptions being Kallisto and Sailfish, which show higher
accuracy, and eXpress, which shows relatively lower accuracy.

\begin{table}
\begin{tabular}[c]{@{}lrrrr@{}}
\toprule\addlinespace
Method & A & B & C & D
\\\addlinespace
\midrule
Isolator & \textbf{0.903} & \textbf{0.908} & \textbf{0.890} &
\textbf{0.884}
\\\addlinespace
eXpress & 0.901 & 0.902 & 0.884 & 0.874
\\\addlinespace
Cufflinks & 0.900 & 0.901 & 0.865 & 0.872
\\\addlinespace
Salmon & 0.899 & 0.897 & 0.882 & 0.871
\\\addlinespace
RSEM/ML & 0.899 & 0.896 & 0.881 & 0.870
\\\addlinespace
Kallisto & 0.893 & 0.885 & 0.874 & 0.860
\\\addlinespace
BitSeq & 0.875 & 0.865 & 0.863 & 0.848
\\\addlinespace
Sailfish & 0.879 & 0.857 & 0.854 & 0.834
\\\addlinespace
RSEM/PM & 0.878 & 0.866 & 0.864 & 0.848
\\\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\caption{Proportionality correlation between gene level quantification
of 806 genes using TaqMan qPCR and RNA-Seq quantification.}
\end{table}

\begin{table}\begin{tabular}[c]{@{}lrrrr@{}}
\toprule\addlinespace
Method & A & B & C & D
\\\addlinespace
\midrule
Isolator & \textbf{0.878} & \textbf{0.866} & \textbf{0.839} &
\textbf{0.852}
\\\addlinespace
Cufflinks & 0.870 & 0.856 & 0.799 & 0.841
\\\addlinespace
eXpress & 0.870 & 0.855 & 0.829 & 0.840
\\\addlinespace
Salmon & 0.866 & 0.852 & 0.826 & 0.836
\\\addlinespace
RSEM/ML & 0.865 & 0.851 & 0.825 & 0.835
\\\addlinespace
BitSeq & 0.840 & 0.821 & 0.802 & 0.813
\\\addlinespace
Kallisto & 0.858 & 0.840 & 0.817 & 0.826
\\\addlinespace
Sailfish & 0.844 & 0.814 & 0.797 & 0.802
\\\addlinespace
RSEM/PM & 0.840 & 0.822 & 0.803 & 0.811
\\\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\caption{Proportionality correlation between gene level quantification
of 18353 genes using PrimePCR qPCR and RNA-Seq quantification.}
\end{table}

\begin{table}\begin{tabular}[c]{@{}lrrrr@{}}
\toprule\addlinespace
Method & A & B & C & D
\\\addlinespace
\midrule
Isolator & \textbf{0.979} & \textbf{0.978} & \textbf{0.981} &
\textbf{0.982}
\\\addlinespace
Salmon & 0.976 & 0.975 & 0.978 & 0.979
\\\addlinespace
Kallisto & 0.972 & 0.972 & 0.973 & 0.976
\\\addlinespace
Sailfish & 0.970 & 0.969 & 0.969 & 0.972
\\\addlinespace
Cufflinks & 0.967 & 0.969 & 0.972 & 0.974
\\\addlinespace
RSEM/PM & 0.943 & 0.949 & 0.944 & 0.949
\\\addlinespace
RSEM/ML & 0.941 & 0.948 & 0.945 & 0.951
\\\addlinespace
BitSeq & 0.940 & 0.949 & 0.943 & 0.946
\\\addlinespace
eXpress & 0.931 & 0.939 & 0.935 & 0.942
\\\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\caption{Proportionality correlation between known proportions of 92
ERCC spike-in controls and RNA-Seq quantification.}
\end{table}

\subsection{Estimation consistency}\label{estimation-consistency}

The four SEQC samples consist of two commercially available reference
RNA samples, labeled $A$ and $B$, and two synthetic samples formed by
mixing $A$ and $B$ in specific proportions. Sample $C$ is composed of
75\% $A$ and 25\% $B$, while sample $D$ is 25\% $A$ and 75\% $B$. This
design allows us to directly measure expression consistency. If we
estimate transcript abundance for each sample producing estimates
$a, b, c, d$, then for these estimates to be in agreement we should
expect $c$ to be approximated by $0.25a + 0.75b$, and similarly $d$ by
$0.75a + 0.25b$. In this manner we measured consistency for each method
using proportionality correlation (Table TODO).

Results from RSEM are particularly informative. We generated maximum
likeilood and posterior mean estimates using the same aligned reads.
Merely switching the estimate from the former to the latter increased
the correlation from 0.922 to 0.968. This stands in contrast to
comparisons to qPCR which suggest reduced accuracy of posterior mean
estimates, demonstrating that consistency and accuracy, while related,
are two separate axis of comparison.

Estimates made at the gene level (Supplementary Table TODO) agree with
these results, but are of uniformly higher correlation and show a much
smaller gap between posterior mean and maximum likelihood estimate.
Posterior mean methods (Isolator, BitSeq, RSEM/PM) again show the
highest correlation and are in very close agreement.

\begin{table}\begin{tabular}[c]{@{}lrr@{}}
\toprule\addlinespace
Method & $c$ vs $0.75a + 0.25b$ & $d$ vs $0.25a + 0.75b$
\\\addlinespace
\midrule
Isolator & \textbf{0.975} & \textbf{0.975}
\\\addlinespace
BitSeq & 0.967 & 0.967
\\\addlinespace
RSEM/PM & 0.968 & 0.967
\\\addlinespace
Sailfish & 0.932 & 0.925
\\\addlinespace
RSEM/ML & 0.922 & 0.919
\\\addlinespace
Salmon & 0.916 & 0.914
\\\addlinespace
Kallisto & 0.907 & 0.902
\\\addlinespace
eXpress & 0.903 & 0.899
\\\addlinespace
Cufflinks & 0.870 & 0.916
\\\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\caption{Proportionality correlation between transcript-level estimates
for the mixed samples C and D and weighted averages of estimates for A
and B, corresponding to the mixture proportions for C and D.}
\end{table}

\begin{table}\begin{tabular}[c]{@{}lrr@{}}
\toprule\addlinespace
Method & $c$ vs $0.75a + 0.25b$ & $d$ vs $0.25a + 0.75b$
\\\addlinespace
\midrule
Isolator & \textbf{0.995} & \textbf{0.995}
\\\addlinespace
BitSeq & 0.994 & 0.994
\\\addlinespace
RSEM/PM & 0.994 & 0.993
\\\addlinespace
RSEM/ML & 0.987 & 0.987
\\\addlinespace
Salmon & 0.986 & 0.986
\\\addlinespace
eXpress & 0.983 & 0.983
\\\addlinespace
Sailfish & 0.978 & 0.976
\\\addlinespace
Kallisto & 0.972 & 0.970
\\\addlinespace
Cufflinks & 0.941 & 0.982
\\\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\caption{Proportionality correlation between gene-level estimates for
the mixed samples C and D and weighted averages of estimates for A and
B, corresponding to the mixture proportions for C and D. TODO: move this
table to the supplement.}
\end{table}

\subsection{Batch effects}\label{batch-effects}

\begin{figure*}
\includegraphics[width=\textwidth]{batch-effect-figure.pdf}
\caption{
\textbf{a} A heatmap showing pairwise proportionality correlation between the sample
sampled sequenced on two flowcells each at five sites. Flowcells are numbered 1
or 2 and sequencing sites are abbreviated with three letter codes: Australian
Genome Research Facility (AGR), Beijing Genome Institute (BGI), Cornell
University (CNL), Mayo Clinic (MAY), and Novartis (NVS). \textbf{b} The absolute
change in correlation induced by enabling bias correction on methods that
support it. For clarity this plot excludes points for BitSeq estimates of "MAY
2", as bias correction has an extreme detrimental effect on these.
}
\end{figure*}

We further examined the question of consistency by comparing the same
samples sequenced on different flow cells and at different sites, again
using data from the SEQC. We used 10 samples from the larger SEQC
dataset, each constituting a single flowcell lane from two separate
flowcells sequenced at five different sites, all on Illumina HiSeq 2000
instruments, and compared pairwise agreement (Figure TODO).

These correlations are somewhat smaller than those in Table TODO,
largely because of shallower sequencing: there five flowcell lanes were
combined, totaling \textasciitilde{}25 millions reads per sample, here
only one lane or \textasciitilde{}5 million reads are used.
Nevertheless, we again posterior mean methods with showing much higher
agreement between pairs.

When accounting for batch effects, bias correction methods can have a
significant impact. To examine the efficacy of bias correction, we
disabled bias correction functionality on those methods that support it,
repeated the experiment, and measured the change in pairwise correlation
(Figure TODO). We see that bias correction is largely beneficial for
each method,

Isolator, Salmon, and Cufflinks show similar improvements with bias
correction, though in a few cases Cufflinks slightly decreases
agreement. Kallisto show a consistent, but slight improvement. BitSeq's
bias correction sometimes has a very positive effect, but other times
negative or disastrous effect. Estimates from the ``MAY 2'' samples in
particular had far worse agreement with other samples with bias
correction enabled.

While this benchmark is informative, it should be considered a lower
bound on the batch effects and bias: these samples were identically
prepared and sequenced. The fact that technical effects is indicative
that RNA-Seq is not so consistent that batch effects can be under any
circumstanced ignored.

\subsection{Accuracy in simulated
data}\label{accuracy-in-simulated-data}

To demonstrate Isolator's capacity to produce more accurate estimates by
modeling entire experiments, we simulated a simple RNA-Seq experiment
consisting of two conditions each with three replicated. Expression
values were generated from a two-component log-normal model with
parameters fit to Cufflinks estimates of data from Kuppusamy et al
(TODO: cite). We generated simulated RNA-Seq reads using rlsim (TODO:
cite).

In this benchmark, Isolator significantly improves on existing methods.
It has the unique advantage of sharing information between replicates.
With the assumption that genes tend to be similarly spliced between
replicates, we are able to more effectively resolve transcript
expression in complex loci. Kallisto, Salmon, RSEM/ML, and Cufflinks all
perform very similarly. While comparisons to qPCR showed eXpress and
Cufflinks with a significant advantage, effective bias correction is
less important in simulated data, plausibly explaining the difference.
Although rlsim models some forms of bias, technical effects in RNA-Seq
are not entirely understood, so these effects are likely understated.

\begin{table}\begin{tabular}[c]{@{}ll@{}}
\toprule\addlinespace
Method & Correlation
\\\addlinespace
\midrule
Isolator & 0.919
\\\addlinespace
Kallisto & 0.887
\\\addlinespace
Salmon & 0.886
\\\addlinespace
RSEM/ML & 0.881
\\\addlinespace
Cufflinks & 0.881
\\\addlinespace
eXpress & 0.825
\\\addlinespace
Sailfish & 0.816
\\\addlinespace
RSEM/PM & 0.806
\\\addlinespace
BitSeq & 0.796
\\\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\caption{Proportionality correlation between ground truth and estimates
produced by each method on simulated RNA-Seq.}
\end{table}

\subsection{Agreement between two sequencing
technologies}\label{agreement-between-two-sequencing-technologies}

RNA-Seq reads become more informative with length and quantity. We
compared data from the same sample sequenced twice, once with 300pb
paired-end reads on a MiSeq sequencer and again with 100bp paired-end
reads using a HiSeq 2000, yielding approximately 4.5 million and 11
million reads, respectively. To elucidate the effect of deeper
sequencing, we sampled without replacement smaller subsets of the 11
million HiSeq 2000 reads and measured correlation between estimates
these subsamples with estimates from the MiSeq data (Figure TODO).

Isolator show the highest correlation, but only when sequencing depth
exceeds several million reads, where it overtakes eXpress by a small
margin. The outliers in the this test are BitSeq and RSEM/PM. The former
shows extremely high correlation with small numbers of reads, and
contrary to all other methods, correlation decreases with deeper
sequencing after 1 million reads. When posterior mean estimates are
generated from a model using a uninformative prior, as is the case with
BitSeq, transcripts with few or no reads may produce estimates that are
highly influenced by transcript length, which plausibly explains this
phenomenon. Yet, RSEM/PM does increase in correlation monotonically with
sequencing depth, though the correlation remains very low. Even using
all 11 million reads, RSEM/PM had a correlation of only 0.567.

\begin{figure}
\includegraphics[width=0.5\textwidth]{rhemac-figure.pdf}
\end{figure}

\subsection{Finding monotonic differential
splicing}\label{finding-monotonic-differential-splicing}

By modeling the entire RNA-Seq experiment and saving samples generated
during MCMC sampling, Isolator is uniquely able to compute posterior
probabilities corresponding to arbitrarily complex questions, within the
confines of the model. As a case study, we previously used Isolator to
analyze splicing dynamics in human cardiomyocyte cells during
maturation.

The experiment consisted of a number of conditions at different time
points so did not easily lend itself to analysis using pairwise
hypothesis tests. We instead looked for genes showed a change in
splicing that was consistent between mature and immature cells by
computing a probability of ``monotonic splicing'', or specifically that
an observed change in splicing occurs in a consistent direction

TODO: talk about validation.

\section{Discussion}\label{discussion}

To assess the suitability of Isolator we have compared it to large
number of alternative methods using a wide variety of benchmarks, with a
particular focus on estimation consistency or stability. Data generated
by the SEQC is ideally suited to examine the question consistency. Our
data agrees with the results published by the SEQC, who compared
Cufflinks and BitSeq and found BitSeq to produce more consistent
estimates. Here we control for alignment method by using the same
alignments for every method (excluding the alignment-free methods
Salmon, Sailfish, and Kallisto), and compare wider variety methods. In
this broader context it becomes clear that this separation is a
consequence of the choice in estimators. Unmoderated maximum likelihood
methods, though often accurate, frequently demonstrate lower consistency
than methods that use MCMC sampling and report posterior means or
medians. Conversely, most methods that report posterior mean estimates
show seemingly poor accuracy.

Much of the cause of the reduced accuracy of the MCMC based models is
driven by low-expression transcripts. A transcript with no reads has a
maximum likelihood expression of zero, but likelihood is positive over a
range of non-zero values, so the posterior mean must necessarily also be
positive. With a flat, or ``non-informative'' prior expression values
for these low-expression transcripts are often quite high and highly
determined by transcript length. Isolator avoids this problem by using a
model with informative priors: namely encoding assumptions that most
transcripts have low expression and that expression and splicing tend to
be similar between samples. In this way, it is able to avoid what seems
like a tradeoff and produce estimates that are both consistent and
accurate.

Fully Bayesian methods like the one presented here possess a degree of
subjectivity that sometimes give researchers pause. Although we have
found the results from Isolator to be insensitive to precise values
given to hyperparameters, they are chosen in advance and with a degree
of arbitrariness. What is often ignored is that the alternative, methods
based on unregularized maximum likelihood estimates, in practice often
necessitate a more insidious form of subjectivity: contriving cutoffs to
filter out unreliable point estimates of low expression genes or
transcripts. While the priors in Bayesian methods are purely explicit,
the means by which data was filtered during an analysis is often a form
of off-the-books subjectivity--critical to the results, yet unmentioned
or only alluded to in manuscripts.

\section{Online Methods}\label{online-methods}

\subsection{Hierarchical Modeling of RNA-Seq
experiments}\label{hierarchical-modeling-of-rna-seq-experiments}

To model RNA-Seq experiments we use a hierarchical model consisting of
three levels: samples or replicates, conditions, and the experiment. Any
number of conditions, and any number of replicates within those
conditions. We model condition-wise and experiment-wise expression for
each transcript using a Gamma distribution, parameterized by mean and
shape. Shape, which capture biological variability, are shared across
conditions.

We further parameterize the model to capture splicing. Despite the
compositional nature, for efficiency purposes we use independent Normal
distributions the model splicing rates at the condition and experiment
level, where a splicing rate is an isoforms proportion relative to total
expression of the its gene.

\subsection{Efficient Sampling}\label{efficient-sampling}

The isoform quantification or deconvolution problem is most easily
thought of as an additive mixture model in which the component
distributions (the isoforms) are known, but the mixture coefficients
(isoform abundances) must be inferred. The standard approach to this
problem is ether expectation maximization (EM) if the maximum likelihood
solution is desired, or Gibbs sampling if, as we are, estimating the
posterior distribution. We have taken a different approach and instead
rely on slice sampling.

The trade-offs between slice sampling and Gibbs sampling for the isoform
quantification problem are subtle. Subsequent samples drawn from a slice
sampler are generally less autocorrelated than those drawn from a Gibbs
sampler. As a result, fewer samples need to be generated to adequately
explore a distribution. However, each of these samples is more
computationally expensive to compute, so one might compensate for
autocorrelation in a Gibbs sampler by drawing more samples in less time.
It is not obvious then that one approach is inherently more efficient
than the other.

Because the likelihood function in this problem is simple, and we sample
from the posterior probability directly, without introducing latent
variables, we are afforded some optimization opportunities not available
in a Gibbs sampler. Specifically, we implement the likelihood function
using SIMD (single instruction, multiple data) instructions, which are
available nearly all CPUs from the last decade. This, in combination
with some numerical approximations, allow us to compute the likelihood
function over an order of magnitude faster on conventional hardware,
than a naive C implementation.

TODO: point to timing table in supplement

\subsection{Correcting for Bias}\label{correcting-for-bias}

Isolator attempts to model and correct for multiple factors that can
conflate the transcript expression estimates.

Perhaps the most prominent source of nonuniformity is sequence specific
bias present at fragment ends. A probable source of this bias is cDNA
synthesis by random priming, which is part of popular RNA-Seq protocols
such as Illumina's TruSeq kit. We have previously published a model that
successfully accounts for effect by training a sparse graphical model
(TODO: cite).

Beyond this random priming bias, there is residual fragment GC-bias,
which may be an artifact of PCR amplification. We observe that fragments
with very high or very low GC-content are less likely to be observed
than expected, even after accounting for priming bias. We model this by
binning fragments by GC content, computing expected and observed
frequencies, and adjusting by their ratio.

Selection of polyadenylated transcripts is a common step mRNA-Seq used
to avoid sequencing introns, partially degraded transcripts, and
ribosomal RNA. This can cause enrichment of reads at the 3' end of
transcripts if only partial transcripts are captured. To model the
effect we fit a one parameter model in which a transcript is truncated
at any position with probability $p$. The probability of observing a
fragment at a position $k$ of length $\ell$, counting from the 3' end,
is then proportional to the probability that the transcript was
\emph{not} truncated before $k + \ell$, which is $(1-p)^{k + \ell}$.
When fit, $p$ is typically quite small, on the order of $10^{-5}$, in
which case this correction has little effect on transcripts shorter that
several kilobases.

Lastly, subtle implications arise from random fragmentation steps in
many protocols. Existing statistical models (see (Pachter 2011) for a
review), assume fragments are sampled uniformly at random from a
transcript. However, this does not exactly match the implications of
random fragmentation in which, under an ideal model, \emph{breakpoints}
rather than fragments are introduced uniformly at random. For a fragment
to be observed it must pass size selection and have fallen between two
breakpoints, or one breakpoint and the end of the transcript. Since the
ends of a transcript act as fixed breakpoints, the result is some
enrichment of fragments at either end of the transcript. We compute this
enrichment exactly for various transcript length, and use interpolation
to approximate th effect.

\section{References}\label{references}

Lovell, David, Vera Pawlowsky-Glahn, Juan Jos{é} Egozcue, Samuel
Marguerat, and J{ü}rg B{ä}hler. 2015. ``Proportionality: a Valid
Alternative to Correlation for Relative Data.'' \emph{PLoS Computational
Biology} 11 (3) (March): e1004075.

Pachter, Lior. 2011. ``Models for Transcript Quantification from
RNA-Seq.'' \emph{arXiv.Org} (April).
\url{http://arxiv.org/abs/1104.3889v2}.

\end{document}

### 2015-08-05 paper.tex
More revision. Start adding junk to supplement.

@@
 \usepackage{hyperref}
+\usepackage[numbers]{natbib}
+\usepackage{xr}
 
-%\setromanfont{Adagio_Serif Light}
-%\setsansfont[BoldFont={Adagio_Sans}]{Adagio_Sans Light}
-%\allsectionsfont{\normalfont\sffamily\bfseries}
+\externaldocument{supplement}
+
+\setromanfont{Adagio_Serif Light}
+\setsansfont[BoldFont={Adagio_Sans}]{Adagio_Sans Light}
+\allsectionsfont{\normalfont\sffamily\bfseries}
 
@@
     isoform-level mRNA quantification, short reads have limitations when
-    resolving complex and similar sets of isoforms. As a result, estimates
-    of isoform abundance, compared to those at the gene level, carry far
-    more uncertainty. When confronted with this uncertainty, commonly used
-    methods produce estimates that are unstable; small perturbations in the
-    data often produce dramatically different results. We introduce a new
-    method, Isolator, which analyzes all samples in an experiment in unison
-    using a simple Bayesian hierarchical model. Combined with aggressive bias
-    correction, it produces estimates that are simultaneously accurate and
+    resolving complex or similar sets of isoforms. As a result, estimates of
+    isoform abundance carry far more uncertainty than those made at the gene
+    level. When confronted with this uncertainty, commonly used methods produce
+    estimates that are unstable; small perturbations in the data often produce
+    dramatically different results confounding downstream analysis. We introduce
+    a new method, Isolator, which analyzes all samples in an experiment in
+    unison using a simple Bayesian hierarchical model. Combined with aggressive
+    bias correction, it produces estimates that are simultaneously accurate and
     stable. In a comprehensive comparison of accuracy and stability, we show
-    that this property is unique to Isolator. We further demonstrate that
-    the approach of modeling an entire experiment enables new analyses by
-    examining splicing monotonicity across several time points in the
-    development of human cardiomyocyte cells.
+    that this property is unique to Isolator. We further demonstrate that the
+    approach of modeling an entire experiment enables new analyses by examining
+    splicing monotonicity across several time points in the development of human
+    cardiomyocyte cells.
 \end{abstract}
@@
 
-Since the introduction of RNA-Seq it has rapidly become a preferred
-method of studying gene expression, having proved to be reliable,
-reproducible, and increasingly affordable. In principle, RNA-Seq enables
-measurements of expression at a higher resolution than has previously
-been possible or practical at the same scale. Rather than an often
-vaguely defined notion of a gene, expression of specific isoform or
-exons can be measured. In practice this promise is difficult to realize.
-There are no standardized names or symbols for isoforms, annotations are
-often incomplete or in disagreement, and methods designed to detect
-changes in splicing rarely concur (TODO: cite).
+Since its introduction, RNA-Seq has rapidly become a preferred method of
+studying gene expression, having proved to be reliable, reproducible, and
+increasingly affordable. In principle, RNA-Seq enables measurements of
+expression at a scale and resolution than has previously been impossible or
+impractical. Rather than relying on an often vaguely defined notion of a gene,
+expression of specific isoforms or exons can be measured. In practice this
+promise is difficult to realize.  There are no standardized names or symbols for
+isoforms, annotations are often incomplete or in disagreement, and methods
+designed to detect changes in splicing rarely concur \cite{Liu:2014dn}.
 
-Because a short read is often compatible with many isoforms, the signal
-produced from sequencing must be deconvoluted, requiring a certain level
-of algorithmic sophistication. This problem was tackled early on and
-with great success by Cufflinks, and indeed many of the methods
-developed in the interim have followed from the same basic model of
-RNA-Seq in which transcripts are represented as particular distributions
-over possible reads, and the problem is then to infer mixing
-coefficients (relative expression values) inducing a mixture model that
-best explains the data.
+Most RNA-Seq procotocols produce short reads (typically 30 to 150 nucleotides).
+Because a short read is often compatible with many isoforms, the signal produced
+from sequencing must be deconvoluted, requiring a certain level of algorithmic
+sophistication. This problem was tackled early on and with great success by
+Cufflinks \cite{Trapnell:2010kd}, and indeed many of the methods developed in
+the interim have followed from the same basic model of RNA-Seq in which
+transcripts are represented as particular distributions over possible reads, and
+the problem is then to infer mixing coefficients (relative expression values)
+inducing a mixture model that best explains the data \cite{Pachter:2011wm}.
 
@@
 
-An efficient MCMC algorithm is used to generate samples over the
-parameters of this model, which are saved in an HDF5 based format. The
-output from the sampler can then be processed to produce point
-estimates, credible intervals, posterior probabilities, and diagnostics.
-Though designed to be run on multiple samples concurrently, Isolator can
-be run as a conventional tool on individual samples.
+An efficient Markov chain Monte Carlo algorithm is used to generate samples over
+the parameters of this model, which are saved in an HDF5 \cite{hdf5} based
+format. The output from the sampler can then be processed to produce point
+estimates, credible intervals, posterior probabilities, and diagnostics. Though
+designed to be run on multiple samples concurrently, Isolator can be run as a
+conventional tool on individual samples.
 
@@
 estimates towards a baseline of no change, producing more conservative
-estimates of effect size than one would obtain by using, for example, a
+estimates of effect size than would be obtained by using, for example, a
 flat Dirichlet prior. This feature of the model is particularly
-important for robust treatment of low-expression genes. With few reads
+important for robust treatment of low-expression transcripts. With few reads
 there is little power to determine splicing patterns. Considering the
@@
 
-In addition to approach to we attempt to aggressively correct for
-technical effects. The naive model of RNA-Seq assumes fragments are
-sampled uniformly at random from transcripts in proportion to their
-abundance. Several have observed that this uniformity assumption is far
-from an accurate description of real sequencing data (TODO: cite).
-Though severe at the nucleotide resolution, gene level expression
-estimates are only moderately affected by this bias, as it tends to
-average out over long transcripts. When one considers expression at the
-isoform level, attempting to correct for this nonuniformity becomes
-vital, as the difference between two isoforms is often only a few
-nucleotides.
+In addition to this hierarchical model, we attempt to aggressively correct for
+technical effects. The naive model of RNA-Seq assumes fragments are sampled
+uniformly at random from transcripts in proportion to their abundance. Several
+have observed that this uniformity assumption is far from an accurate
+description of real sequencing data \cite{Hansen:2010if}.  Though severe at the
+nucleotide resolution, gene level expression estimates are only moderately
+affected by this bias, as it tends to average out over long transcripts. When
+one considers expression at the isoform level, attempting to correct for this
+nonuniformity becomes vital, as the difference between two isoforms is often
+only a few nucleotides.
 
-We previously developed an efficient and versatile model to account for
-sequence specific bias that commonly occur at the ends of fragments
-(TODO: cite). In Isolator, we supplemented this with methods accounting
-for 3' bias, fragment GC-content bias, and positional bias caused by
-end-effects of fragmentation.
+We previously developed an efficient and versatile model to account for sequence
+specific bias that commonly occurs at the ends of fragments \cite{Jones:2012hg}
+In Isolator, we supplemented this with methods accounting for 3' bias, fragment
+GC-content bias, and positional bias caused by end-effects of fragmentation.
 
@@
 
-\section{Results}\label{results}
+\section{Results}
 
-The convention for evaluating the accuracy of RNA-Seq analysis methods
-has consisted chiefly of comparing point estimate accuracy in
-simulations, supplemented with comparisons to qPCR or other
-technologies. Tuning methods exclusively to maximize accuracy on
-simulated data creates the risk of solving only an idealized
-mathematical model of RNA-Seq while disregarding noise, bias, and
+The convention for evaluating the accuracy of RNA-Seq analysis methods has
+consisted chiefly of comparing point estimate accuracy in simulations,
+supplemented with comparisons to qPCR or other technologies. Tuning methods
+exclusively to maximize accuracy on simulated data risks solving only an
+idealized mathematical model of RNA-Seq while disregarding noise, bias, and
 nonuniformity. As no gold standard for whole-transcriptome isoform-level
-estimation exists, we adopt multi-faceted approach, with a focus
-measuring the stability or consistency of estimates, along with the
-accuracy.
+estimation exists, we adopt multi-faceted approach, with a focus on measuring
+the stability or consistency of estimates, along with accuracy.
 
-Comparisons of gene or transcript expression is sensitive to how one
+Comparisons of gene or transcript expression are sensitive to how one
 chooses to measure similarity or distance. Because gene expression
-typically varies across several orders of magnitude, measurements like
+typically varies across many orders of magnitude, measurements like
 the Pearson correlation, $L_1$ or $L_2$ distance, or root-mean-square
@@
 transcripts which have no, or very low expression. These issues are
-discussed in depth by Lovell et al (Lovell et al. 2015), who propose
+discussed in depth by Lovell et al \cite{Lovell:2015il}, who propose
 measurements of proportionality as a more meaningful metric. We adopt
@@
 varies from -1 to 1, with 1 indicating perfectly proportionality, and -1
-perfect reciprocality. Zeros are accounted for by adding 0.1 to all TPM
-values.
+perfect reciprocality. Zeros are accounted for by converting all measurements to
+transcripts per million (TPM) and adding 0.1.
 
@@
 estimate the relative abundance a set of known transcripts. These
-include Cufflinks, RSEM, eXpress, BitSeq, Sailfish, Salmon, BitSeq, and
-Kallisto. RSEM was used to produce both maximum likelihood and posterior
-mean estimates, which we label ``RSEM/ML'' and ``RSEM/PM'' respectively.
+include, non-exhaustively, Cufflinks \cite{Trapnell:2010kd},
+RSEM \cite{Li:2011cb},
+eXpress \cite{Roberts:2012dh},
+BitSeq \cite{Glaus:2012ek},
+Sailfish \cite{Patro:2014jd},
+Salmon (a refinement of Sailfish currently in development), and
+Kallisto \cite{Bray:2015uj}.
+
+RSEM was used to produce both maximum likelihood and posterior
+mean estimates, which we label ``RSEM/ML'' and ``RSEM/PM'', respectively.
 This is not an exhaustive account of such methods, but does represent a
@@
 
-\subsection{Agreement with qPCR and spike-in
-controls}\label{agreement-with-qpcr-and-spike-in-controls}
+\subsection{Agreement with qPCR and spike-in controls}
 
@@
 
-We use data generated by the SEQC project, consisting of four reference
-samples (labeled A, B, C, D). The RNA-seq methods were run using
-approximately 25 million reads per sample, obtaining estimates that were
-then compared to qPCR in Tables TODO and TODO, and the known ERCC
-spike-in mixtures in Table TODO. In each of these we find that Isolator
-produces the highest correlation of the methods compared, though
-sometimes by negligible margin. Despite two different technologies, and
-two different sets of genes, the qPCR benchmarks agree closely, ranking
-the methods nearly identically. Cufflinks, eXpress, Salmon, and RSEM/ML
-typically perform similarly while BitSeq, Kallisto, Sailfish, and
-RSEM/PM trail slightly. Results are similar with ERCC spike-in controls,
-the notable exceptions being Kallisto and Sailfish, which show higher
-accuracy, and eXpress, which shows relatively lower accuracy.
+We use data generated by the Sequencing Quality Control Consortium (SEQC)
+\cite{Consortium:2014kc}, consisting of four reference samples (labeled A, B, C,
+and D). The RNA-seq methods were run using approximately 25 million reads per
+sample, obtaining estimates that were then compared to qPCR in Tables
+\ref{table:taqman} and \ref{table:primepcr}, and the known ERCC spike-in
+mixtures in Table \ref{table:ercc}. In each of these we find that Isolator
+produces the highest correlation of the methods compared, though sometimes by a
+negligible margin. Despite two different technologies, and two different sets of
+genes, the qPCR benchmarks agree closely, ranking the methods nearly
+identically. Cufflinks, eXpress, Salmon, and RSEM/ML typically perform similarly
+to each other, ith BitSeq, Kallisto, Sailfish, and RSEM/PM trailing slightly.
+Results are similar with ERCC spike-in controls, the notable exceptions being
+Kallisto and Sailfish, which show higher accuracy, and eXpress, which shows
+lower accuracy compared to their relative rankings in the qPCr benchmarks.
 
@@
 of 806 genes using TaqMan qPCR and RNA-Seq quantification.}
+\label{table:taqman}
 \end{table}
@@
 of 18353 genes using PrimePCR qPCR and RNA-Seq quantification.}
+\label{table:primepcr}
 \end{table}
@@
 ERCC spike-in controls and RNA-Seq quantification.}
+\label{table:ercc}
 \end{table}
@@
 
-The four SEQC samples consist of two commercially available reference
-RNA samples, labeled $A$ and $B$, and two synthetic samples formed by
-mixing $A$ and $B$ in specific proportions. Sample $C$ is composed of
-75\% $A$ and 25\% $B$, while sample $D$ is 25\% $A$ and 75\% $B$. This
-design allows us to directly measure expression consistency. If we
-estimate transcript abundance for each sample producing estimates
-$a, b, c, d$, then for these estimates to be in agreement we should
-expect $c$ to be approximated by $0.25a + 0.75b$, and similarly $d$ by
-$0.75a + 0.25b$. In this manner we measured consistency for each method
-using proportionality correlation (Table TODO).
+The four SEQC samples consist of two commercially available reference RNA
+samples, labeled $A$ and $B$, and two synthetic samples formed by mixing $A$ and
+$B$ in specific proportions. Sample $C$ is composed of 75\% $A$ and 25\% $B$,
+while sample $D$ is 25\% $A$ and 75\% $B$. This design allows us to directly
+measure expression consistency. If we estimate transcript abundance for each
+sample, producing estimates $a, b, c, d$, then for these estimates to be in
+agreement we should expect $c$ to be approximated by $0.25a + 0.75b$, and
+similarly $d$ by $0.75a + 0.25b$. To capture the notion of conistency, we
+measured the closeness of this approximation using proportionality correlation
+(Table \ref{table:seqctranscripts}).
 
@@
 
-Estimates made at the gene level (Supplementary Table TODO) agree with
-these results, but are of uniformly higher correlation and show a much
-smaller gap between posterior mean and maximum likelihood estimate.
-Posterior mean methods (Isolator, BitSeq, RSEM/PM) again show the
-highest correlation and are in very close agreement.
+Estimates made at the gene level (Supplementary Table \ref{table:seqcgenes})
+agree with these results, but are of uniformly higher correlation and show a
+much smaller gap between posterior mean and maximum likelihood estimate.
+Methods reporting posterior mean estimates (Isolator, BitSeq, RSEM/PM) again
+show the highest correlation and are in very close agreement.
 
@@
 and B, corresponding to the mixture proportions for C and D.}
-\end{table}
-
-\begin{table}\begin{tabular}[c]{@{}lrr@{}}
-\toprule\addlinespace
-Method & $c$ vs $0.75a + 0.25b$ & $d$ vs $0.25a + 0.75b$
-\\\addlinespace
-\midrule
-Isolator & \textbf{0.995} & \textbf{0.995}
-\\\addlinespace
-BitSeq & 0.994 & 0.994
-\\\addlinespace
-RSEM/PM & 0.994 & 0.993
-\\\addlinespace
-RSEM/ML & 0.987 & 0.987
-\\\addlinespace
-Salmon & 0.986 & 0.986
-\\\addlinespace
-eXpress & 0.983 & 0.983
-\\\addlinespace
-Sailfish & 0.978 & 0.976
-\\\addlinespace
-Kallisto & 0.972 & 0.970
-\\\addlinespace
-Cufflinks & 0.941 & 0.982
-\\\addlinespace
-\bottomrule
-\addlinespace
-\end{tabular}
-\caption{Proportionality correlation between gene-level estimates for
-the mixed samples C and D and weighted averages of estimates for A and
-B, corresponding to the mixture proportions for C and D. TODO: move this
-table to the supplement.}
+\label{table:seqctranscripts}
 \end{table}
@@
 \caption{
-\textbf{a} A heatmap showing pairwise proportionality correlation between the sample
-sampled sequenced on two flowcells each at five sites. Flowcells are numbered 1
+\textbf{a} A heatmap showing pairwise proportionality correlation between the
+samples sequenced on two flowcells each at five sites. Flowcells are numbered 1
 or 2 and sequencing sites are abbreviated with three letter codes: Australian
@@
 University (CNL), Mayo Clinic (MAY), and Novartis (NVS). \textbf{b} The absolute
-change in correlation induced by enabling bias correction on methods that
+change in correlation induced by enabling bias correction for methods that
 support it. For clarity this plot excludes points for BitSeq estimates of "MAY
-2", as bias correction has an extreme detrimental effect on these.
-}
+2", as bias correction has an extreme detrimental effect on these.}
+\label{fig:batcheffects}
 \end{figure*}
@@
 using data from the SEQC. We used 10 samples from the larger SEQC
-dataset, each constituting a single flowcell lane from two separate
+dataset, each consisting of a single lane from two separate
 flowcells sequenced at five different sites, all on Illumina HiSeq 2000
-instruments, and compared pairwise agreement (Figure TODO).
+instruments, and compared pairwise agreement between estimates from these
+samples (Figure \ref{fig:batcheffects}a).
 
-These correlations are somewhat smaller than those in Table TODO,
+These correlations are somewhat smaller than those in Table
+\ref{table:seqctranscripts},
 largely because of shallower sequencing: there five flowcell lanes were
@@
 only one lane or \textasciitilde{}5 million reads are used.
-Nevertheless, we again posterior mean methods with showing much higher
+Nevertheless, we again see posterior mean estimates with significantly higher
 agreement between pairs.
@@
 When accounting for batch effects, bias correction methods can have a
-significant impact. To examine the efficacy of bias correction, we
-disabled bias correction functionality on those methods that support it,
-repeated the experiment, and measured the change in pairwise correlation
-(Figure TODO). We see that bias correction is largely beneficial for
-each method,
+significant impact. To examine the efficacy of bias correction, we disabled this
+functionality on those methods that support it, repeated the experiment, and
+measured the change in pairwise correlation (Figure \ref{fig:batcheffects}b). We
+see that bias correction is largely beneficial for each method. Isolator,
+Salmon, and Cufflinks show similar improvements with bias correction, though in
+a few cases Cufflinks slightly decreases agreement. Kallisto show a consistent,
+but very slight improvement. BitSeq's bias correction sometimes has a very positive
+effect, but other times negative or disastrous effect. Estimates from the ``MAY
+2'' samples in particular had far worse agreement with other samples with bias
+correction enabled. Here bias correction has dramatic and detrimental effects on
+some number of transcripts, driving the correlation to negative numbers.
 
-Isolator, Salmon, and Cufflinks show similar improvements with bias
-correction, though in a few cases Cufflinks slightly decreases
-agreement. Kallisto show a consistent, but slight improvement. BitSeq's
-bias correction sometimes has a very positive effect, but other times
-negative or disastrous effect. Estimates from the ``MAY 2'' samples in
-particular had far worse agreement with other samples with bias
-correction enabled.
+While this benchmark is informative, it should be considered a lower bound on
+the batch effects and bias: these samples were identically prepared and
+sequenced. The fact that technical effects were so clearly present even under
+such controlled circumstances is indicative that RNA-Seq is not so reproducible
+that batch effects can be, under any circumstanced ignored.
 
-While this benchmark is informative, it should be considered a lower
-bound on the batch effects and bias: these samples were identically
-prepared and sequenced. The fact that technical effects is indicative
-that RNA-Seq is not so consistent that batch effects can be under any
-circumstanced ignored.
+\subsection{Accuracy in simulated data}\label{accuracy-in-simulated-data}
 
-\subsection{Accuracy in simulated
-data}\label{accuracy-in-simulated-data}
-
-To demonstrate Isolator's capacity to produce more accurate estimates by
-modeling entire experiments, we simulated a simple RNA-Seq experiment
-consisting of two conditions each with three replicated. Expression
-values were generated from a two-component log-normal model with
-parameters fit to Cufflinks estimates of data from Kuppusamy et al
-(TODO: cite). We generated simulated RNA-Seq reads using rlsim (TODO:
-cite).
+To more directly demonstrate Isolator's capacity to produce more accurate
+estimates by modeling entire experiments, we simulated a simple RNA-Seq
+experiment consisting of two conditions each with three replicated. Expression
+values were generated from a two-component log-normal model with parameters fit
+to Cufflinks estimates of data from Kuppusamy et al \cite{Kuppusamy:2015ey}. We
+generated 10 million, 100nt paired-end simulated RNA-Seq reads per sample using
+rlsim \cite{Sipos:2013tz}.
 
@@
 Although rlsim models some forms of bias, technical effects in RNA-Seq
-are not entirely understood, so these effects are likely understated.
+are not entirely understood, so these are likely understated in its model.
 
@@
 produced by each method on simulated RNA-Seq.}
+\label{table:rlsim}
 \end{table}
@@
 RNA-Seq reads become more informative with length and quantity. We
-compared data from the same sample sequenced twice, once with 300pb
-paired-end reads on a MiSeq sequencer and again with 100bp paired-end
+compared data from the same sample sequenced twice, once with 300nt
+paired-end reads on a MiSeq sequencer and again with 100nt paired-end
 reads using a HiSeq 2000, yielding approximately 4.5 million and 11
@@
 sequencing, we sampled without replacement smaller subsets of the 11
-million HiSeq 2000 reads and measured correlation between estimates
-these subsamples with estimates from the MiSeq data (Figure TODO).
+million HiSeq 2000 reads and measured correlation between estimates made using
+these subsamples with estimates from the MiSeq data (Figure \ref{fig:rhemac}).
 
-Isolator show the highest correlation, but only when sequencing depth
+Isolator shows the highest correlation, but only when sequencing depth
 exceeds several million reads, where it overtakes eXpress by a small
 margin. The outliers in the this test are BitSeq and RSEM/PM. The former
-shows extremely high correlation with small numbers of reads, and
-contrary to all other methods, correlation decreases with deeper
+shows extremely high correlation with small numbers of reads, but
+contrary to all other methods, correlation rapidly decreases with deeper
 sequencing after 1 million reads. When posterior mean estimates are
@@
 \includegraphics[width=0.5\textwidth]{rhemac-figure.pdf}
+\caption{Proportionality correlation between estimates from 4.5 million 300nt
+MiSeq reads and progressively larger numbers of HiSeq 2000 reads.}
+\label{fig:rhemac}
 \end{figure}
 
-\subsection{Finding monotonic differential
-splicing}\label{finding-monotonic-differential-splicing}
+\subsection{Finding monotonic differential splicing}\label{finding-monotonic-differential-splicing}
 
 By modeling the entire RNA-Seq experiment and saving samples generated
-during MCMC sampling, Isolator is uniquely able to compute posterior
+during sampling, Isolator is uniquely able to compute posterior
 probabilities corresponding to arbitrarily complex questions, within the
@@
 The experiment consisted of a number of conditions at different time
-points so did not easily lend itself to analysis using pairwise
+points so it did not easily lend itself to analysis using pairwise
 hypothesis tests. We instead looked for genes showed a change in
@@
 computing a probability of ``monotonic splicing'', or specifically that
-an observed change in splicing occurs in a consistent direction
+an observed change in splicing occurs in a consistent direction.
 
@@
 
-To assess the suitability of Isolator we have compared it to large
-number of alternative methods using a wide variety of benchmarks, with a
-particular focus on estimation consistency or stability. Data generated
-by the SEQC is ideally suited to examine the question consistency. Our
-data agrees with the results published by the SEQC, who compared
-Cufflinks and BitSeq and found BitSeq to produce more consistent
-estimates. Here we control for alignment method by using the same
-alignments for every method (excluding the alignment-free methods
-Salmon, Sailfish, and Kallisto), and compare wider variety methods. In
-this broader context it becomes clear that this separation is a
-consequence of the choice in estimators. Unmoderated maximum likelihood
-methods, though often accurate, frequently demonstrate lower consistency
-than methods that use MCMC sampling and report posterior means or
-medians. Conversely, most methods that report posterior mean estimates
-show seemingly poor accuracy.
+To assess the suitability of Isolator we have compared it to large number of
+alternative methods using a wide variety of benchmarks, with a particular focus
+on estimation consistency or stability. Data generated by the SEQC is ideally
+suited to examine the question consistency, and indeed our data agrees with
+their results. The SEQC compared Cufflinks and BitSeq and found BitSeq to
+produce more consistent estimates. Here we control for alignment method by using
+the same alignments for every method (excluding the alignment-free methods
+Salmon, Sailfish, and Kallisto), and compare wider variety methods. In this
+broader context it becomes clear that this separation is a consequence of the
+choice in estimators. Unmoderated maximum likelihood methods, though often
+accurate, frequently demonstrate lower consistency than methods that use MCMC
+sampling and report posterior means or medians. This should be unsurprising.
+Because of the ambiguity of sequencing data, the likelihood functions for
+relative expression are often near plateaus. Maximum likelihood estimates are
+thus are precarious compared to posterior mean estimates. Neither estimate is
+inherently more correct, and robust analysis should make use of the full
+posterior distribution, but we argue that posterior mean or median estimates
+are more useful in practice because of this property.
 
-Much of the cause of the reduced accuracy of the MCMC based models is
-driven by low-expression transcripts. A transcript with no reads has a
-maximum likelihood expression of zero, but likelihood is positive over a
-range of non-zero values, so the posterior mean must necessarily also be
-positive. With a flat, or ``non-informative'' prior expression values
-for these low-expression transcripts are often quite high and highly
-determined by transcript length. Isolator avoids this problem by using a
-model with informative priors: namely encoding assumptions that most
-transcripts have low expression and that expression and splicing tend to
-be similar between samples. In this way, it is able to avoid what seems
-like a tradeoff and produce estimates that are both consistent and
+Most methods that report posterior mean estimates show seemingly poor accuracy.
+Much of the cause of the seemingly reduced accuracy of the MCMC based models is
+driven by low-expression transcripts. A transcript with no reads has a maximum
+likelihood expression of zero, but likelihood is positive over a range of
+non-zero values, so the posterior mean must necessarily also be positive. With
+a flat, or ``non-informative'' prior expression values for these low-expression
+transcripts are often quite high and highly determined by transcript length.
+Isolator avoids this problem by using a model with informative priors: namely
+encoding assumptions that most transcripts have low expression and that
+expression and splicing tend to be similar between samples. In this way, it is
+able to avoid what seems like a tradeoff and produce estimates that are both
+consistent and
 accurate.
@@
 successfully accounts for effect by training a sparse graphical model
-(TODO: cite).
+\cite{Jones:2012hg}.
 
@@
 
-Lastly, subtle implications arise from random fragmentation steps in
-many protocols. Existing statistical models (see (Pachter 2011) for a
-review), assume fragments are sampled uniformly at random from a
-transcript. However, this does not exactly match the implications of
-random fragmentation in which, under an ideal model, \emph{breakpoints}
-rather than fragments are introduced uniformly at random. For a fragment
-to be observed it must pass size selection and have fallen between two
-breakpoints, or one breakpoint and the end of the transcript. Since the
-ends of a transcript act as fixed breakpoints, the result is some
-enrichment of fragments at either end of the transcript. We compute this
-enrichment exactly for various transcript length, and use interpolation
-to approximate th effect.
+Lastly, subtle implications arise from random fragmentation steps in many
+protocols. Existing statistical models assume fragments are sampled uniformly
+at random from a transcript. However, this does not exactly match the
+implications of random fragmentation in which, under an ideal model,
+\emph{breakpoints} rather than fragments are introduced uniformly at random. For
+a fragment to be observed it must pass size selection and have fallen between
+two breakpoints, or one breakpoint and the end of the transcript. Since the ends
+of a transcript act as fixed breakpoints, the result is some enrichment of
+fragments at either end of the transcript. We compute this enrichment exactly
+for various transcript length, and use interpolation to approximate th effect.
 
-\section{References}\label{references}
-
-Lovell, David, Vera Pawlowsky-Glahn, Juan Jos{é} Egozcue, Samuel
-Marguerat, and J{ü}rg B{ä}hler. 2015. ``Proportionality: a Valid
-Alternative to Correlation for Relative Data.'' \emph{PLoS Computational
-Biology} 11 (3) (March): e1004075.
-
-Pachter, Lior. 2011. ``Models for Transcript Quantification from
-RNA-Seq.'' \emph{arXiv.Org} (April).
-\url{http://arxiv.org/abs/1104.3889v2}.
+\bibliographystyle{plainnat}
+\bibliography{references,extra-references}
 

### 2015-08-05 supplement.tex
More revision. Start adding junk to supplement.

\documentclass{article}
\usepackage[T1]{fontenc}
\usepackage[cm-default]{fontspec}
\usepackage{lmodern}

\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{sectsty}
\usepackage{ctable}
\usepackage{hyperref}
\usepackage[numbers]{natbib}

\begin{document}

% TODO:
%   * Explain that every SEQC sample was put in the same experiment for
%     isolator.
%   * Move gene-expression consistency table here.
%   *

\section{Bias Correction}\label{bias-correction}

Isolator attempts to model and correct for multiple factors that can
conflate the transcript expression estimates.

\subsection{3' bias}\label{bias}

Selection of polyadenylated transcripts is a common step mRNA-Seq used
to avoid sequencing introns, partially degraded transcripts, and
ribosomal RNA.

To model the effect we fit a one parameter model in which a transcript
is truncated at any position with probability $p$. The probability of
observing a fragment at a position $k$ of length $\ell$, counting from
the 3' end, is then proportional to the probability that the transcript
was \emph{not} truncated before $k + \ell$, which is $(1-p)^{k + \ell}$.
When fit, $p$ is typically quite small, on the order of $10^{-5}$, in
which case this correction has little effect on transcripts shorter that
several kilobases.

\subsection{Fragmentation effects}\label{fragmentation-effects}

Popular RNA-Seq protocols such as Illumina's TruSeq typically involve a
random fragmentation step. Subtle implications arise from this process.

Existing statisticaly models (see {[}@Pachter:2011wm{]} for a review),
assume fragments are sampled uniformly at random from a transcript. This
does not exactly match the implications of random fragmentation in
which, under an ideal model, \emph{breakpoints} rather than fragments
are introduced uniformly at random. For a fragment to be observed it
must pass size selection and have fallen between two breakpoints, or one
breakpoint and the end of the transcript. Since the ends of a transcript
act as fixed breakpoints, the result isAsome enrichment of fragments at
either end of the transcript.

\subsection{Fragment GC content}\label{fragment-gc-content}

\section{Analysis}

All reads were aligned using the STAR aligner \cite{Dobin:2013fg}. RNA-Seq
methods vary in their expected input: some work on raw reads, others require
either alignments relative to genome or relative to transcript sequences. STAR
allows the same alignments to be output in transcript and genome coordinates, so
we were able to use the same alignments for each method, excluding the alignment
free methods. Quantification was performed against version 76 of the Ensembl
annotations and revision 38 of the human genome sequence.

\begin{table}
\begin{tabular}[c]{ll}
    STAR & 2.4.0k \\
    Cufflinks & 2.2.1 \\
    BitSeq & 0.7.0 \\
    RSEM & 1.2.19 \\
    Sailfish & 0.6.3 \\
    eXpress & 1.5.1 \\
    Kallisto & 0.42 \\
    Salmon & 0.4.1
\end{tabular}
\caption{
    Version of software used in the analysis.
}
\end{table}

\begin{table}\begin{tabular}[c]{@{}lrr@{}}
\toprule\addlinespace
Method & $c$ vs $0.75a + 0.25b$ & $d$ vs $0.25a + 0.75b$
\\\addlinespace
\midrule
Isolator & \textbf{0.995} & \textbf{0.995}
\\\addlinespace
BitSeq & 0.994 & 0.994
\\\addlinespace
RSEM/PM & 0.994 & 0.993
\\\addlinespace
RSEM/ML & 0.987 & 0.987
\\\addlinespace
Salmon & 0.986 & 0.986
\\\addlinespace
eXpress & 0.983 & 0.983
\\\addlinespace
Sailfish & 0.978 & 0.976
\\\addlinespace
Kallisto & 0.972 & 0.970
\\\addlinespace
Cufflinks & 0.941 & 0.982
\\\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\caption{Proportionality correlation between gene-level estimates for
the mixed samples C and D and weighted averages of estimates for A and
B, corresponding to the mixture proportions for C and D. TODO: move this
table to the supplement.}
\label{table:seqcgenes}
\end{table}

\section{Analysis of cardiomyocyte data}

Different methods often produce very different estimates of isoform abundance.
To demonstrate the effect this has on downstream analysis, we analyzed the
cardiomyocyte data from \cite{Kuppusamy:2015ey} using each method compared in
the main paper.

\subsection{qPCR Validation}

Total RNAs for 2 independent HFA and HFV samples were extracted using RNAeasy
mini kit (Qiagen) (6). Two independent HAH total RNA samples were purchased
commercially from Clontech and Agilent. Total RNAs for 3 independent replicates
of Day 20 cardiomyocytes (derived from human embryonic stem cells) were
extracted with miRNeasy mini kit (Qiagen). All RNA samples were further DNAse
treated with DNA-free kit (Ambion) and reverse transcription reaction was
carried out using Omniscript RT kit (Qiagen). Gene specific primers were
designed targeting the exon-exon junction for specific exons of interest using
the freely available software APE. The primer sequences were as follows:
TNNT2-F- CAGAGACCATGTCTGACATAGA and R-CTGCCTCCTCCTGCTCGT; OBSL1-F-
CCTGTACGCTGGTACAAGGA and R-CTCTTCCACGCTGACAAGGA; LTBP4-F- CAGCCCAAAAAGTGTGCAGG
and R-CTGTCTCTTCCTCACGCTGG; OSBPL1A-F- TCACAGCTCTGCTCAACAGG and
R-ACATTGTTTATGGGCCCGGT; ACOT9-F-CCACATTCATGAAGCATGTTCA and
R-CCTACTATCTCCCGCAACT, IMMT-F-TGCACTTTCAGAAGAAGCATCC and R-TTTTCCTGTTGTGCAAGGCG;
PALM-F-TGAAGGCAGCCATGTACTCG and R-GGACCACTTTGGTCTCGTCC

\begin{figure}
\includegraphics[width=0.5\textwidth]{analysis/cardio/ACOT9-comparison.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/IMMT-comparison.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/LTBP4-comparison.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/OBSL1-comparison.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/OSBPL1A-comparison.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/PALM-comparison.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/TNNT2-comparison.pdf}
\caption{Point estimates of fold change of exon expression from qPCR and and
RNA-Seq using various methods. Exon expression for RNA-Seq is computed by
estimating transcript expression, then summing the expression of all transcripts
that share a parcitular exon.}
\end{figure}

\subsection{Principal Component Analysis}

A simple analysis often peformed on gene expression data is principal component
analysis (PCA). Due to large disagreements in isoform abundance estimates, each
method produces qualitatively different results. There is no ground truth for
this data, but the results from Isolator in some ways fit our understanding of
the data more closely than the other methods. Uniquely in the Isolator data,
fetal atrium and fetal ventricle are grouped very closely and clearly separated
from the H7 day20 data.

In most methods the two adult samples are quite heterogenous. The exceptions are
Kallisto and eXpress. While those both show closer agreement between the adult
samples, the fetal samples are far more spread out.

Though Isolator uses a model with informative priors, it is important to note
that the fact that fetal atrium and ventricle tissue in particular are similar
is not something that is encoded is this model.

\begin{figure}
\includegraphics[width=0.5\textwidth]{analysis/cardio/isolator-pca.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/salmon-pca.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/sailfish-pca.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/kallisto-pca.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/express-pca.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/cufflinks-pca.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/pm-rsem-pca.pdf}
\includegraphics[width=0.5\textwidth]{analysis/cardio/ml-rsem-pca.pdf}
\caption{The first two principal components from principal component analysis
applied to transcript expression estimates using a variety of methods.}
\end{figure}

\end{document}

### 2015-08-08 paper.tex
Stuff on qpcr validation.

@@
 agreement we should expect $c$ to be approximated by $0.25a + 0.75b$, and
-similarly $d$ by $0.75a + 0.25b$. To capture the notion of conistency, we
+similarly $d$ by $0.75a + 0.25b$. To capture the notion of consistency, we
 measured the closeness of this approximation using proportionality correlation
@@
 Results from RSEM are particularly informative. We generated maximum
-likeilood and posterior mean estimates using the same aligned reads.
+likelihood and posterior mean estimates using the same aligned reads.
 Merely switching the estimate from the former to the latter increased
@@
 
-By modeling the entire RNA-Seq experiment and saving samples generated
-during sampling, Isolator is uniquely able to compute posterior
-probabilities corresponding to arbitrarily complex questions, within the
-confines of the model. As a case study, we previously used Isolator to
-analyze splicing dynamics in human cardiomyocyte cells during
-maturation.
+Splicing has been known to play an important role in devoloping hearts. Specific
+splicing changes during maturation have been shown to be conserved across
+chicken and mouse, and influenced by expression of CELF and MBNL
+\cite{Kalsotra:2008ke}. Recently, stress-induced switching between two mutually
+exclusive isoforms of KHK has been shown to be driver of pathological cardiac
+hypertrophy \cite{Mirtschink:2015iq}.
 
-The experiment consisted of a number of conditions at different time
-points so it did not easily lend itself to analysis using pairwise
-hypothesis tests. We instead looked for genes showed a change in
-splicing that was consistent between mature and immature cells by
-computing a probability of ``monotonic splicing'', or specifically that
-an observed change in splicing occurs in a consistent direction.
+We previously used Isolator to analyze splicing dynamics in human cardiomyocyte
+cells during maturation \cite{Kuppusamy:2015ey}. By modeling the entire RNA-Seq
+experiment and saving samples generated during sampling, Isolator is uniquely
+able to compute posterior probabilities corresponding to arbitrarily complex
+questions, within the confines of the model. As this experiment consisted of
+a number of samples from a number of conditions and time points, it did not
+easily lend itself to analysis using pairwise tests.
 
-TODO: talk about validation.
+We instead searched for differential splicing events by computing the
+probability of ``monotonic splicing``, which we defined as a splicing change
+that that occurs is consistently observed as occuring in the same direction
+between immature (H7 cells at day 20 and at day 30 infected with an empty
+vector) and mature samples (H7 1 Year, adult heart, and fetal atrium and
+ventricle). To compute the posterior probability of monotonic splicing, we
+simply count the number of samples from the model in which this pattern is
+observed, and divide by the total number of samples generated (350, in this
+case).
+
+This approach identified 272 probable (posterior probability > 0.5) differential
+splicing events. We chose seven high probability candidates (ACOT9, IMMT, LTBP4,
+OBSL1, OSBPL1A, PALM, and TNNT2) to validate with qPCR (Supplementary Section
+\ref{cardioqpcr}), by measuring the expression of specific exons predicted a
+change in splicing. Five of these candidates (ACOT9, LTBP4, OBSL1, OSBPL1A, and
+TNNT2) showed a clear gradient between H7 day20, fetal, and adult tissue.
+
+We analyzed the same data with the other seven methods. Exon-level fold change
+estimates generally agree between RNA-Seq methods, though far less so between
+RNA-Seq and qPCR, which could be caused by poor gene annotations or yet
+unidentified technical effects. Principal component analysis using transcript
+expression estimates also demonstrates some revealing differences (Supplementary
+Section \ref{cardiopca}).
 

### 2015-10-25 paper.tex
Figure tweaks.

@@
 Isolator & \textbf{0.903} & \textbf{0.908} & \textbf{0.890} &
-\textbf{0.884}
+\textbf{0.883}
 \\\addlinespace
@@
 Isolator & \textbf{0.878} & \textbf{0.866} & \textbf{0.839} &
-\textbf{0.852}
+\textbf{0.851}
 \\\addlinespace
@@
 \caption{Proportionality correlation between gene level quantification
-of 18353 genes using PrimePCR qPCR and RNA-Seq quantification.}
+of 18,353 genes using PrimePCR qPCR and RNA-Seq quantification.}
 \label{table:primepcr}
@@
 \midrule
-Isolator & \textbf{0.979} & \textbf{0.978} & \textbf{0.981} &
-\textbf{0.982}
+Isolator & \textbf{0.980} & \textbf{0.979} & \textbf{0.981} &
+\textbf{0.983}
 \\\addlinespace
@@
 \midrule
-Isolator & \textbf{0.975} & \textbf{0.975}
+Isolator & \textbf{0.978} & \textbf{0.978}
 \\\addlinespace
@@
 \caption{Proportionality correlation between transcript-level estimates
-for the mixed samples C and D and weighted averages of estimates for A
-and B, corresponding to the mixture proportions for C and D.}
+for the mixed SEQC samples C and D and weighted averages of estimates for A
+and B, corresponding to the mixture proportions for C and D. Sample C consists of
+$3/4$ A and $1/4$ B and sample D is $1/4$ A and $3/4$ B. Consistent estimates of
+transcript expression in A, B, C, and D should agree with these mixture
+proportions.}
 \label{table:seqctranscripts}
@@
 \caption{Proportionality correlation between ground truth and estimates
-produced by each method on simulated RNA-Seq.}
+produced by each method using simulated RNA-Seq reads.}
 \label{table:rlsim}
@@
 \caption{Proportionality correlation between estimates from 4.5 million 300nt
-MiSeq reads and progressively larger numbers of HiSeq 2000 reads.}
+MiSeq reads and progressively larger numbers of HiSeq 2000 reads. For clarity,
+correlations for RSEM/PM are excluded as they were much lower than the other
+methods.}
 \label{fig:rhemac}
@@
 function over an order of magnitude faster on conventional hardware,
-than a naive C implementation.
-
-TODO: point to timing table in supplement
+than a more straightforward implementation. Run time for Isolator is generally
+significantly less than that of other sampling based implementations, though
+alignment-free, maximum-likelihood methods like Salmon and Kallisto remain the
+most efficient methods (Supplementary Figure \ref{fig:runtime})
 

### 2015-10-25 supplement.tex
Stuff on qpcr validation.
Figure tweaks.

@@
 %     isolator.
-%   * Move gene-expression consistency table here.
-%   *
 
@@
 \begin{table}
+\begin{center}
 \begin{tabular}[c]{ll}
-    STAR & 2.4.0k \\
-    Cufflinks & 2.2.1 \\
-    BitSeq & 0.7.0 \\
-    RSEM & 1.2.19 \\
-    Sailfish & 0.6.3 \\
-    eXpress & 1.5.1 \\
-    Kallisto & 0.42 \\
-    Salmon & 0.4.1
+\toprule\addlinespace
+Method & Version \\
+\midrule
+STAR & 2.4.0k \\\addlinespace
+Cufflinks & 2.2.1 \\\addlinespace
+BitSeq & 0.7.0 \\\addlinespace
+RSEM & 1.2.19 \\\addlinespace
+Sailfish & 0.6.3 \\\addlinespace
+eXpress & 1.5.1 \\\addlinespace
+Kallisto & 0.42 \\\addlinespace
+Salmon & 0.4.1 \\\addlinespace
+\bottomrule
+\addlinespace
 \end{tabular}
-\caption{
-    Version of software used in the analysis.
-}
+\caption{Versions of software used in the analysis.}
+\end{center}
 \end{table}
 
-\begin{table}\begin{tabular}[c]{@{}lrr@{}}
+\begin{table}
+\begin{center}
+\begin{tabular}[c]{@{}lrr@{}}
 \toprule\addlinespace
@@
 table to the supplement.}
+\end{center}
 \label{table:seqcgenes}
 \end{table}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=0.7\textwidth]{analysis/seqc/runtime.pdf}
+\caption{Run time needed to process the SEQC data presented in the Results
+sections. All methods were run a Google Compute Engine instance backed by four
+Intel Xeon cores and 52GB of a memory.}
+\end{center}
+\label{fig:runtime}
+\end{figure}
 
@@
 
-\subsection{qPCR Validation}
+\subsection{qPCR Validation}\label{cardioqpcr}
 
@@
 the freely available software APE. The primer sequences were as follows:
-TNNT2-F- CAGAGACCATGTCTGACATAGA and R-CTGCCTCCTCCTGCTCGT; OBSL1-F-
-CCTGTACGCTGGTACAAGGA and R-CTCTTCCACGCTGACAAGGA; LTBP4-F- CAGCCCAAAAAGTGTGCAGG
-and R-CTGTCTCTTCCTCACGCTGG; OSBPL1A-F- TCACAGCTCTGCTCAACAGG and
-R-ACATTGTTTATGGGCCCGGT; ACOT9-F-CCACATTCATGAAGCATGTTCA and
-R-CCTACTATCTCCCGCAACT, IMMT-F-TGCACTTTCAGAAGAAGCATCC and R-TTTTCCTGTTGTGCAAGGCG;
-PALM-F-TGAAGGCAGCCATGTACTCG and R-GGACCACTTTGGTCTCGTCC
+
+\begin{center}
+\begin{tabular}{rl}
+TNNT2   & F-CAGAGACCATGTCTGACATAGA \\
+        & R-CTGCCTCCTCCTGCTCGTx \\
+OBSL1   & F-CCTGTACGCTGGTACAAGGA \\
+        & R-CTCTTCCACGCTGACAAGGA \\
+LTBP4   & F-CAGCCCAAAAAGTGTGCAGG \\
+        & R-CTGTCTCTTCCTCACGCTGG \\
+OSBPL1A & F-TCACAGCTCTGCTCAACAGG \\
+        & R-ACATTGTTTATGGGCCCGGT \\
+ACOT9   & F-CCACATTCATGAAGCATGTTCA \\
+        & R-CCTACTATCTCCCGCAACT \\
+IMMT    & F-TGCACTTTCAGAAGAAGCATCC \\
+        & R-TTTTCCTGTTGTGCAAGGCG \\
+PALM    & F-TGAAGGCAGCCATGTACTCG \\
+        & R-GGACCACTTTGGTCTCGTCC
+\end{tabular}
+\end{center}
 
@@
 that share a parcitular exon.}
+\label{fig:cardiofoldchange}
 \end{figure}
 
-\subsection{Principal Component Analysis}
+\subsection{Principal Component Analysis}\label{cardiopca}
 
@@
 applied to transcript expression estimates using a variety of methods.}
+\label{fig:cardiopca}
 \end{figure}

### 2015-11-07 paper.tex
Combine some tables into one figure, etc.

@@
 \usepackage{hyperref}
+\usepackage{subcaption}
 \usepackage[numbers]{natbib}
@@
 
-\externaldocument{supplement}
+% TODO: wtf does this break things?
+%\externaldocument{supplement}
 
@@
 \title{Isolator: accurate and consistent analysis of RNA-Seq experiments}
+
+% Possible alternative titles? Larry suggests putting 'isoform' in the name.
+% 
+% accurate and consistent isoform quantification
+%
+%
+%
+
 \author{Daniel C. Jones \and Kavitha T. Kuppusamy \and Xinxia Peng \and Michael G. Katze \and Hannele Ruohola-Baker \and Walter L. Ruzzo}
@@
 and D). The RNA-seq methods were run using approximately 25 million reads per
-sample, obtaining estimates that were then compared to qPCR in Tables
+sample, obtaining estimates that were then compared to qPCR in Figures
 \ref{table:taqman} and \ref{table:primepcr}, and the known ERCC spike-in
-mixtures in Table \ref{table:ercc}. In each of these we find that Isolator
+mixtures in Figure \ref{table:ercc}. In each of these we find that Isolator
 produces the highest correlation of the methods compared, though sometimes by a
@@
 
-\begin{table}
+\begin{figure*}
+\begin{subtable}{0.5\linewidth}
+\caption{Comparison of estimates to TaqMan qPCR}
 \begin{tabular}[c]{@{}lrrrr@{}}
@@
 \end{tabular}
-\caption{Proportionality correlation between gene level quantification
-of 806 genes using TaqMan qPCR and RNA-Seq quantification.}
 \label{table:taqman}
-\end{table}
-
-\begin{table}\begin{tabular}[c]{@{}lrrrr@{}}
+\end{subtable}
+\begin{subtable}{0.5\linewidth}
+\caption{Comparison of estimates to PrimePCR qPCR}
+\begin{tabular}[c]{@{}lrrrr@{}}
 \toprule\addlinespace
@@
 \end{tabular}
-\caption{Proportionality correlation between gene level quantification
-of 18,353 genes using PrimePCR qPCR and RNA-Seq quantification.}
 \label{table:primepcr}
-\end{table}
+\end{subtable}
 
-\begin{table}\begin{tabular}[c]{@{}lrrrr@{}}
+\begin{subtable}{0.5\linewidth}
+\caption{Comparison of estimates to ERCC spike-in controls}
+\begin{tabular}[c]{@{}lrrrr@{}}
 \toprule\addlinespace
@@
 \end{tabular}
-\caption{Proportionality correlation between known proportions of 92
-ERCC spike-in controls and RNA-Seq quantification.}
 \label{table:ercc}
-\end{table}
-
-\subsection{Estimation consistency}\label{estimation-consistency}
-
-The four SEQC samples consist of two commercially available reference RNA
-samples, labeled $A$ and $B$, and two synthetic samples formed by mixing $A$ and
-$B$ in specific proportions. Sample $C$ is composed of 75\% $A$ and 25\% $B$,
-while sample $D$ is 25\% $A$ and 75\% $B$. This design allows us to directly
-measure expression consistency. If we estimate transcript abundance for each
-sample, producing estimates $a, b, c, d$, then for these estimates to be in
-agreement we should expect $c$ to be approximated by $0.25a + 0.75b$, and
-similarly $d$ by $0.75a + 0.25b$. To capture the notion of consistency, we
-measured the closeness of this approximation using proportionality correlation
-(Table \ref{table:seqctranscripts}).
-
-Results from RSEM are particularly informative. We generated maximum
-likelihood and posterior mean estimates using the same aligned reads.
-Merely switching the estimate from the former to the latter increased
-the correlation from 0.922 to 0.968. This stands in contrast to
-comparisons to qPCR which suggest reduced accuracy of posterior mean
-estimates, demonstrating that consistency and accuracy, while related,
-are two separate axis of comparison.
-
-Estimates made at the gene level (Supplementary Table \ref{table:seqcgenes})
-agree with these results, but are of uniformly higher correlation and show a
-much smaller gap between posterior mean and maximum likelihood estimate.
-Methods reporting posterior mean estimates (Isolator, BitSeq, RSEM/PM) again
-show the highest correlation and are in very close agreement.
-
-\begin{table}\begin{tabular}[c]{@{}lrr@{}}
+\end{subtable}
+\begin{subtable}{0.5\linewidth}
+\caption{Consistency of SEQC sample estimates}
+\begin{tabular}[c]{@{}lrr@{}}
 \toprule\addlinespace
@@
 \end{tabular}
-\caption{Proportionality correlation between transcript-level estimates
+\label{table:seqctranscripts}
+\end{subtable}
+
+\caption{
+Benchmarks measuring the accuracy and consistency of estimates made from 
+Sequencing Quality Control (SEQC) dataset using a variety of methods.
+All numbers are reported in proportionality correlation.
+(a) correlation between gene level quantification
+of 806 genes using TaqMan qPCR and RNA-Seq quantification.
+(b) correlation between gene level quantification
+of 18,353 genes using PrimePCR qPCR and RNA-Seq quantification.
+(c) correlation between known proportions of 92
+ERCC spike-in controls and RNA-Seq quantification.
+(d) correlation between transcript-level estimates
 for the mixed SEQC samples C and D and weighted averages of estimates for A
@@
 proportions.}
-\label{table:seqctranscripts}
-\end{table}
+\end{figure*}
+
+\subsection{Estimation consistency}\label{estimation-consistency}
+
+The four SEQC samples consist of two commercially available reference RNA
+samples, labeled $A$ and $B$, and two synthetic samples formed by mixing $A$ and
+$B$ in specific proportions. Sample $C$ is composed of 75\% $A$ and 25\% $B$,
+while sample $D$ is 25\% $A$ and 75\% $B$. This design allows us to directly
+measure expression consistency. If we estimate transcript abundance for each
+sample, producing estimates $a, b, c, d$, then for these estimates to be in
+agreement we should expect $c$ to be approximated by $0.25a + 0.75b$, and
+similarly $d$ by $0.75a + 0.25b$. To capture the notion of consistency, we
+measured the closeness of this approximation using proportionality correlation
+(Figure \ref{table:seqctranscripts}).
+
+Results from RSEM are particularly informative. We generated maximum
+likelihood and posterior mean estimates using the same aligned reads.
+Merely switching the estimate from the former to the latter increased
+the correlation from 0.922 to 0.968. This stands in contrast to
+comparisons to qPCR which suggest reduced accuracy of posterior mean
+estimates, demonstrating that consistency and accuracy, while related,
+are two separate axis of comparison.
+
+Estimates made at the gene level (Supplementary Table \ref{table:seqcgenes})
+agree with these results, but are of uniformly higher correlation and show a
+much smaller gap between posterior mean and maximum likelihood estimate.
+Methods reporting posterior mean estimates (Isolator, BitSeq, RSEM/PM) again
+show the highest correlation and are in very close agreement.
 
@@
 
-\begin{table}\begin{tabular}[c]{@{}ll@{}}
+\begin{table}
+\label{table:rlsim}
+\begin{tabular}[c]{@{}ll@{}}
 \toprule\addlinespace
@@
 \midrule
-Isolator & 0.919
+\textbf{Isolator} & \textbf{0.919}
 \\\addlinespace
@@
 \end{tabular}
-\caption{Proportionality correlation between ground truth and estimates
+\caption{
+Proportionality correlation between ground truth and estimates
 produced by each method using simulated RNA-Seq reads.}
-\label{table:rlsim}
 \end{table}
@@
 
-\subsection{Finding monotonic differential splicing}\label{finding-monotonic-differential-splicing}
+\subsection{Case study: finding monotonic differential splicing}\label{finding-monotonic-differential-splicing}
 
@@
 Section \ref{cardiopca}).
+
+\subsection{Case study: novel mechanisms of cell fate specification identified
+by gene splice variants}
+
+TODO: add nathan's text here
 

### 2015-11-07 supplement.tex
Combine some tables into one figure, etc.

@@
 \usepackage[numbers]{natbib}
+\usepackage{subcaption}
 
@@
 
-\begin{figure}
-\includegraphics[width=0.5\textwidth]{analysis/cardio/ACOT9-comparison.pdf}
-\includegraphics[width=0.5\textwidth]{analysis/cardio/IMMT-comparison.pdf}
-\includegraphics[width=0.5\textwidth]{analysis/cardio/LTBP4-comparison.pdf}
-\includegraphics[width=0.5\textwidth]{analysis/cardio/OBSL1-comparison.pdf}
-\includegraphics[width=0.5\textwidth]{analysis/cardio/OSBPL1A-comparison.pdf}
-\includegraphics[width=0.5\textwidth]{analysis/cardio/PALM-comparison.pdf}
-\includegraphics[width=0.5\textwidth]{analysis/cardio/TNNT2-comparison.pdf}
+\begin{figure}[H]
+\includegraphics[width=0.2\textwidth]{analysis/cardio/colorkey.pdf} \\
+\includegraphics[width=\textwidth]{analysis/cardio/ACOT9-comparison.pdf}
+\includegraphics[width=\textwidth]{analysis/cardio/IMMT-comparison.pdf}
+\includegraphics[width=\textwidth]{analysis/cardio/LTBP4-comparison.pdf}
+\end{figure}
+\begin{figure}[H]
+\ContinuedFloat
+\includegraphics[width=\textwidth]{analysis/cardio/OBSL1-comparison.pdf}
+\includegraphics[width=\textwidth]{analysis/cardio/OSBPL1A-comparison.pdf}
+\includegraphics[width=\textwidth]{analysis/cardio/PALM-comparison.pdf}
+\includegraphics[width=\textwidth]{analysis/cardio/TNNT2-comparison.pdf}
 \caption{Point estimates of fold change of exon expression from qPCR and and
@@
 
-\begin{figure}
+\begin{figure}[H]
+\includegraphics[width=0.2\textwidth]{analysis/cardio/colorkey.pdf} \\
 \includegraphics[width=0.5\textwidth]{analysis/cardio/isolator-pca.pdf}
@@
 \includegraphics[width=0.5\textwidth]{analysis/cardio/kallisto-pca.pdf}
+\end{figure}
+\begin{figure}
+\ContinuedFloat
 \includegraphics[width=0.5\textwidth]{analysis/cardio/express-pca.pdf}

### 2016-02-16 paper.tex
Add Nathan's story. Misc improvements.

@@
 \usepackage{xr}
+\usepackage{authblk}
 
@@
 
-\title{Isolator: accurate and consistent analysis of RNA-Seq experiments}
+\title{Isolator: accurate and consistent analysis of isoform-level expression in RNA-Seq experiments}
 
-% Possible alternative titles? Larry suggests putting 'isoform' in the name.
-% 
-% accurate and consistent isoform quantification
-%
-%
-%
+\author[1]{Daniel C. Jones}
+\author[2]{Kavitha T. Kuppusamy}
+\author[3]{Nathan J. Palpant}
+\author[4]{Xinxia Peng}
+\author[4]{Michael G. Katze}
+\author[5]{Charles E. Murry}
+\author[6]{Hannele Ruohola-Baker}
+\author[1,7,8]{Walter L. Ruzzo}
 
-\author{Daniel C. Jones \and Kavitha T. Kuppusamy \and Xinxia Peng \and Michael G. Katze \and Hannele Ruohola-Baker \and Walter L. Ruzzo}
-\date{}
+%\date{}
+
+\affil[1]{Department of Computer Science and Engineering, University of Washington}
+\affil[2]{XXX}
+\affil[3]{XXX}
+\affil[4]{Department of Microbiology, University of Washington}
+\affil[5]{Department of Pathology, University of Washington}
+\affil[6]{Department of Biochemistry, University of Washington}
+\affil[7]{Department of Genome Sciences, University of Washington}
+\affil[8]{Fred Hutchinson Cancer Research Center}
+
+%\affil[1]{Department of Computer Science and Engineering,
+%University of Washington, Seattle, WA 98195-2350, USA}
+%\affil[2]{Department of Genome Sciences, University of Washington, Seattle, WA
+%98195-5065, USA}
+%\affil[3]{Fred Hutchinson Cancer Research Center, Seattle, WA 98109, USA}
+%\affil[4]{Department of Microbiology, University of Washington, Seattle, WA
+%98195-7242, USA}
 
@@
     level. When confronted with this uncertainty, commonly used methods produce
-    estimates that are unstable; small perturbations in the data often produce
-    dramatically different results confounding downstream analysis. We introduce
-    a new method, Isolator, which analyzes all samples in an experiment in
-    unison using a simple Bayesian hierarchical model. Combined with aggressive
-    bias correction, it produces estimates that are simultaneously accurate and
-    stable. In a comprehensive comparison of accuracy and stability, we show
-    that this property is unique to Isolator. We further demonstrate that the
-    approach of modeling an entire experiment enables new analyses by examining
-    splicing monotonicity across several time points in the development of human
+    estimates that are unstable---small perturbations in the data often produce
+    dramatically different results, confounding downstream analysis. We
+    introduce a new method, Isolator, which analyzes all samples in an
+    experiment in unison using a simple Bayesian hierarchical model. Combined
+    with aggressive bias correction, it produces estimates that are
+    simultaneously accurate and stable. In a comprehensive comparison of
+    accuracy and stability, we show that this property is unique to Isolator. We
+    further demonstrate that the approach of modeling an entire experiment
+    enables new analyses, which we demonstrate by examining splicing
+    monotonicity across several time points in the development of human
     cardiomyocyte cells.
@@
 increasingly affordable. In principle, RNA-Seq enables measurements of
-expression at a scale and resolution than has previously been impossible or
-impractical. Rather than relying on an often vaguely defined notion of a gene,
+expression at a scale and resolution than has previously been impractical or
+impossible. Rather than relying on an often vaguely defined notion of a gene,
 expression of specific isoforms or exons can be measured. In practice this
-promise is difficult to realize. There are no standardized names or symbols for
-isoforms, annotations are often incomplete or in disagreement, and methods
-designed to detect changes in splicing rarely concur \cite{Liu:2014dn}.
+promise is difficult to realize, with methods designed to detect changes in
+splicing rarely concuring \cite{Liu:2014dn}.
 
-Most RNA-Seq procotocols produce short reads (typically 30 to 150 nucleotides).
-Because a short read is often compatible with many isoforms, the signal produced
-from sequencing must be deconvoluted, requiring a certain level of algorithmic
-sophistication. This problem was tackled early on and with great success by
-Cufflinks \cite{Trapnell:2010kd}, and indeed many of the methods developed in
-the interim have followed from the same basic model of RNA-Seq in which
-transcripts are represented as particular distributions over possible reads, and
-the problem is then to infer mixing coefficients (relative expression values)
-inducing a mixture model that best explains the data \cite{Pachter:2011wm}.
+Most RNA-Seq procotocols produce short reads (typically 30 to 150 nucleotides,
+often paired-end).  Because a short read is often compatible with many isoforms,
+the signal produced from sequencing must be deconvoluted, requiring a certain
+level of algorithmic sophistication. This problem was tackled early on and with
+great success by Cufflinks \cite{Trapnell:2010kd}, and indeed many of the
+methods developed in the interim have followed from the same basic model of
+RNA-Seq in which transcripts are represented as particular distributions over
+possible reads, and the problem is then to infer mixing coefficients (relative
+expression values) inducing a mixture model that best explains the data
+\cite{Pachter:2011wm}.
 
@@
 it retains the same underlying model of RNA-Seq, we approach the problem
-from the perspective that isoform-level estimates possess a degree
+from the perspective that isoform-level estimates possess a degree of
 uncertainty that renders maximum likelihood point-estimates alone a
@@
 Rather than estimating expression of individual samples, Isolator uses a
-hierarchical model of an entire experiment. In addition to inferring
-expression values for the individual samples, we introduce parameters
-representing condition-wise and experiment-wise expression and splicing,
-as well as per-transcript variance parameters that are shared across
-conditions. As expression studies often use only a small number of
-replicates, shared variance parameters allows us to make more accurate
-estimates of biological variability.
+hierarchical model of an entire experiment, for example, including all samples
+and time points in a time series. In addition to inferring expression values for
+the individual samples, we introduce parameters representing condition-wise and
+experiment-wise expression and splicing, as well as per-transcript variance
+parameters that are shared across conditions. As expression studies often use
+only a small number of replicates, shared variance parameters allows us to make
+more accurate estimates of biological variability.
 
 An efficient Markov chain Monte Carlo algorithm is used to generate samples over
-the parameters of this model, which are saved in an HDF5 \cite{hdf5} based
-format. The output from the sampler can then be processed to produce point
+the parameters of this model, which are saved in an HDF5-based format
+\cite{hdf5}. The output from the sampler can then be processed to produce point
 estimates, credible intervals, posterior probabilities, and diagnostics. Though
@@
 Statistical tests based on these estimates, without considering the full
-posterior distribution, make approximations by either either
+posterior distribution, make approximations by either
 disregarding the relative or compositional nature of the data and
@@
 technical effects. The naive model of RNA-Seq assumes fragments are sampled
-uniformly at random from transcripts in proportion to their abundance. Several
-have observed that this uniformity assumption is far from an accurate
-description of real sequencing data \cite{Hansen:2010if}. Though severe at the
-nucleotide resolution, gene level expression estimates are only moderately
-affected by this bias, as it tends to average out over long transcripts. When
-one considers expression at the isoform level, attempting to correct for this
-nonuniformity becomes vital, as the difference between two isoforms is often
-only a few nucleotides.
+uniformly at random from transcripts in proportion to their abundance. Early on,
+this was shown to be far from an accurate description of real sequencing data
+\cite{Hansen:2010if}. Though severe at nucleotide resolution, gene level
+expression estimates are only moderately affected by this bias, as it tends to
+average out over long transcripts. When one considers expression at the isoform
+level, attempting to correct for this nonuniformity becomes vital, as the
+difference between two isoforms is often only a few nucleotides.
 
 We previously developed an efficient and versatile model to account for sequence
-specific bias that commonly occurs at the ends of fragments \cite{Jones:2012hg}
+specific bias that commonly occurs at the ends of fragments \cite{Jones:2012hg}.
 In Isolator, we supplemented this with methods accounting for 3' bias, fragment
@@
 nonuniformity. As no gold standard for whole-transcriptome isoform-level
-estimation exists, we adopt multi-faceted approach, with a focus on measuring
+estimation exists, we adopt a multifaceted approach, focusing on measuring
 the stability or consistency of estimates, along with accuracy.
 
-Comparisons of gene or transcript expression are sensitive to how one
-chooses to measure similarity or distance. Because gene expression
-typically varies across many orders of magnitude, measurements like
-the Pearson correlation, $L_1$ or $L_2$ distance, or root-mean-square
-error are often dominated by a small subset of the most highly expressed
-genes. Recognizing this, a popular alternative is the Spearman rank
-correlation, which conversely can overemphasize the majority of
-transcripts which have no, or very low expression. These issues are
-discussed in depth by Lovell et al \cite{Lovell:2015il}, who propose
-measurements of proportionality as a more meaningful metric. We adopt
-their ``proportionality correlation'', defined as
-$\text{pcor}(x, y) = 2\text{cov}(\log(x), \log(y)) / (\text{var}(\log(x)) + \text{var}(\log(y)))$.
-Similar to other measures of correlation, proportionality correlation
-varies from -1 to 1, with 1 indicating perfectly proportionality, and -1
-perfect reciprocality. Zeros are accounted for by converting all measurements to
-transcripts per million (TPM) and adding 0.1.
+Comparisons of gene or transcript expression are sensitive to how one chooses to
+measure similarity. Because gene expression typically varies across many orders
+of magnitude, measurements like the Pearson correlation, $L_1$ or $L_2$
+distance, or root-mean-square error are often dominated by the most highly
+expressed genes. Recognizing this, a popular alternative is Spearman's rank
+correlation, which conversely can overemphasize the majority of transcripts,
+which have no, or very low expression. These issues are discussed in depth by
+Lovell et al \cite{Lovell:2015il}, who propose measurements of proportionality
+as a more meaningful metric. We adopt their ``proportionality correlation'',
+defined as $\text{pcor}(x, y) = 2\text{cov}(\log(x), \log(y)) /
+(\text{var}(\log(x)) + \text{var}(\log(y)))$, where is $\text{cov}(x)$ and
+$\text{var}(x)$ give the covariance and variance of $x$, respectively.  Similar
+to other measures of correlation, proportionality correlation varies from -1 to
+1, with 1 indicating perfectly proportionality, and -1 perfect reciprocality.
+Zeros are accounted for by converting all measurements to transcripts per
+million (TPM) and adding 0.1.
+% TODO: add supplementary info about why 0.1 is a reasonable value.
 
@@
 
-RSEM was used to produce both maximum likelihood and posterior
-mean estimates, which we label ``RSEM/ML'' and ``RSEM/PM'', respectively.
-This is not an exhaustive account of such methods, but does represent a
-wide variety of popular approaches.
+RSEM was used to produce both maximum likelihood and posterior mean estimates,
+which we label ``RSEM/ML'' and ``RSEM/PM'', respectively.  This is not an
+exhaustive account of such methods, but does represent a wide variety of popular
+approaches. When possible, the same read alignments were used for each method,
+which notably excludes the alignment-free methods Sailfish, Salmon, and
+Kallisto.
+% TODO: see supplementary section XXX
 
@@
 produces the highest correlation of the methods compared, though sometimes by a
-negligible margin. Despite two different technologies, and two different sets of
+small margin. Despite two different technologies, and two different sets of
 genes, the qPCR benchmarks agree closely, ranking the methods nearly
 identically. Cufflinks, eXpress, Salmon, and RSEM/ML typically perform similarly
-to each other, ith BitSeq, Kallisto, Sailfish, and RSEM/PM trailing slightly.
+to each other, with BitSeq, Kallisto, Sailfish, and RSEM/PM trailing slightly.
 Results are similar with ERCC spike-in controls, the notable exceptions being
 Kallisto and Sailfish, which show higher accuracy, and eXpress, which shows
-lower accuracy compared to their relative rankings in the qPCr benchmarks.
+lower accuracy compared to their relative rankings in the qPCR benchmarks.
 
@@
 
-Results from RSEM are particularly informative. We generated maximum
-likelihood and posterior mean estimates using the same aligned reads.
-Merely switching the estimate from the former to the latter increased
-the correlation from 0.922 to 0.968. This stands in contrast to
-comparisons to qPCR which suggest reduced accuracy of posterior mean
-estimates, demonstrating that consistency and accuracy, while related,
-are two separate axis of comparison.
+Results from RSEM are particularly informative. We generated maximum likelihood
+and posterior mean estimates using the same aligned reads. Merely switching the
+estimate from the former to the latter increased the correlation by
+approximately 5\%. This stands in contrast to comparisons to qPCR which suggest
+reduced accuracy of posterior mean estimates, demonstrating that consistency and
+accuracy, while related, are two separate axis of comparison.
 
@@
 \caption{
-\textbf{a} A heatmap showing pairwise proportionality correlation between the
-samples sequenced on two flowcells each at five sites. Flowcells are numbered 1
-or 2 and sequencing sites are abbreviated with three letter codes: Australian
-Genome Research Facility (AGR), Beijing Genome Institute (BGI), Cornell
-University (CNL), Mayo Clinic (MAY), and Novartis (NVS). \textbf{b} The absolute
-change in correlation induced by enabling bias correction for methods that
-support it. For clarity this plot excludes points for BitSeq estimates of "MAY
-2", as bias correction has an extreme detrimental effect on these.}
+\textbf{a} A heatmap showing pairwise proportionality correlation between
+samples sequenced on two flowcells each at five sites, from centrally prepared
+libraries. Flowcells are numbered arbitrarily 1 or 2 and sequencing sites are
+abbreviated with three letter codes: Australian Genome Research Facility (AGR),
+Beijing Genome Institute (BGI), Cornell University (CNL), Mayo Clinic (MAY), and
+Novartis (NVS). \textbf{b} The absolute change in correlation induced by
+enabling bias correction for methods that support it. For clarity this plot
+excludes points for BitSeq estimates of "MAY 2", as bias correction has an
+extremely detrimental effect on these. Mean improvement in correlation was 0.008
+with Salmon, 0.007 with Cufflinks, 0.006 with Isolator, 0.003 with Kallisto,
+0.002 with eXpress, and -0.164 with BitSeq.
+}
+
 \label{fig:batcheffects}
@@
 These correlations are somewhat smaller than those in Table
-\ref{table:seqctranscripts},
-largely because of shallower sequencing: there five flowcell lanes were
-combined, totaling \textasciitilde{}25 millions reads per sample, here
-only one lane or \textasciitilde{}5 million reads are used.
-Nevertheless, we again see posterior mean estimates with significantly higher
-agreement between pairs.
+\ref{table:seqctranscripts}, largely because of shallower sequencing: in Table
+\ref{table:seqctranscripts} five flowcell lanes were combined, totaling
+\textasciitilde{}25 millions reads per sample, here only one lane
+(\textasciitilde{}5 million reads) was used. Nevertheless, we again see
+posterior mean estimates (as in Isolator, BitSeq, and RSEM/PM) with
+significantly higher agreement between pairs.
 
@@
 a few cases Cufflinks slightly decreases agreement. Kallisto show a consistent,
-but very slight improvement. BitSeq's bias correction sometimes has a very positive
-effect, but other times negative or disastrous effect. Estimates from the ``MAY
-2'' samples in particular had far worse agreement with other samples with bias
-correction enabled. Here bias correction has dramatic and detrimental effects on
-some number of transcripts, driving the correlation to negative numbers.
+but very slight improvement. BitSeq's bias correction sometimes has a very
+positive effect, but other times a negative or even disastrous effect. Estimates
+from the ``MAY 2'' samples in particular had far worse agreement with other
+samples with bias correction enabled. Here bias correction has dramatic and
+detrimental effects on some number of transcripts, driving the correlation to
+negative numbers.
 
 While this benchmark is informative, it should be considered a lower bound on
-the batch effects and bias: these samples were identically prepared and
-sequenced. The fact that technical effects were so clearly present even under
-such controlled circumstances is indicative that RNA-Seq is not so reproducible
-that batch effects can be, under any circumstanced ignored.
+the batch effects and bias found in real experiments: these samples were
+sequenced at different labs, but using the same instrument model (HiSeq 2000)
+and from libraries prepared centrally. The fact that technical effects were so
+clearly present even under such controlled circumstances suggests that
+reproducibility should not be taken for granted in RNA-Seq experiments,
+particularly when it comes to batch effects.
 
@@
 estimates by modeling entire experiments, we simulated a simple RNA-Seq
-experiment consisting of two conditions each with three replicated. Expression
+experiment consisting of two conditions each with three replicates. Expression
 values were generated from a two-component log-normal model with parameters fit
@@
 
-In this benchmark, Isolator significantly improves on existing methods.
-It has the unique advantage of sharing information between replicates.
-With the assumption that genes tend to be similarly spliced between
-replicates, we are able to more effectively resolve transcript
-expression in complex loci. Kallisto, Salmon, RSEM/ML, and Cufflinks all
-perform very similarly. While comparisons to qPCR showed eXpress and
-Cufflinks with a significant advantage, effective bias correction is
-less important in simulated data, plausibly explaining the difference.
-Although rlsim models some forms of bias, technical effects in RNA-Seq
-are not entirely understood, so these are likely understated in its model.
+In this benchmark, Isolator significantly improves on existing methods (Table
+\ref{table:rlsim}). It has the unique advantage of sharing information between
+replicates. With the assumption that genes tend to be similarly spliced between
+replicates, we are able to more effectively resolve transcript expression in
+complex loci. Kallisto, Salmon, RSEM/ML, and Cufflinks all perform very
+similarly. While comparisons to qPCR showed eXpress and Cufflinks with a
+significant advantage over RSEM/ML, Kallisto, and Bitseq, effective bias
+correction is less important in simulated data, plausibly explaining the
+difference. Although rlsim models some forms of bias, technical effects in
+RNA-Seq are not entirely understood, so these are likely understated in its
+model.
 
@@
 Isolator shows the highest correlation, but only when sequencing depth
-exceeds several million reads, where it overtakes eXpress by a small
+exceeds a few million reads (below which would be unusually shallow
+sequencing for an RNA-Seq experiment) where it overtakes eXpress by a small
 margin. The outliers in the this test are BitSeq and RSEM/PM. The former
@@
 
-TODO: add nathan's text here
+\begin{figure*}
+\begin{center}
+\includegraphics[width=0.75\textwidth]{palpant/figure.pdf}
+\end{center}
+\caption{
+\textbf{a}
+Hierarchical clustering of the dominant splice variant for 84
+differentially spliced genes. Blue arrows indicate genes with known function in
+endothelial or heart development. Red arrows indicate genes in the class of
+receptors that bind VEGF.
+\textbf{b}
+Two dimensional principle component analysis in mesodermal progenitor
+populations showing discrete separation of populations entirely based on
+differences in splice variants. 
+\textbf{c}
+Analysis of gene ontology showing enriched categories found in 84 differentially
+spliced genes. (d-e) Heat map analysis 
+\textbf{d}
+Heat map showing expression of annotated isoforms of NRP1.
+\textbf{e}
+Read coverage over isoforms of NRP1.
+\textbf{f}
+Schematic diagram of protein domains encoded by splice variant variants of NRP1.}
+\label{fig:mesodermal}
+\end{figure*}
+
+Very little is known about the role of splice variants in mediating cell fate
+specification even though it has been shown that 92-95\% of genes are
+differentially spliced in a cell or tissue specific manner \cite{Wang:2008ea,
+Pan:2008jq}. In order to identify novel mechanisms of cell fate specification in
+mesoderm development we began by analyzing RNA-seq data to determine differences
+in splice variants between human embryonic stem cell (hESC) derived cardiac
+progenitor cells (CPCs) and two endothelial lineages, hESC-derived endocardial
+endothelial cells (EECs) and hemogenic endothelial cells (HECs) as we have
+described previously \cite{Palpant:2015dv}. 84 genes were identified as
+differentially spliced (defined here as a > 0.3 shift in relative transcript
+abundance with > 0.5 posterior probability) between these populations (Figure
+\ref{fig:mesodermal}a).  Principle component analysis indicated that splice
+variants, independent of gene expression levels, delineate endothelial
+populations relative to the CPCs (Figure \ref{fig:mesodermal}b). Furthermore,
+these data indicate that splice variants alone are sufficient to differentiate
+these closely related lineages.  Gene ontology analysis of the 84 differentially
+spliced genes showed appropriate enrichment for categories associated with
+alternative splicing (Figure \ref{fig:mesodermal}c). Upon closer examination, we
+identified neuropilin-1 (NRP1) as an excellent candidate molecule that showed
+highly distinct expression profiles of splice variants between the cardiac and
+hemogenic endothelial lineages (Figure \ref{fig:mesodermal}d-f).
+
+NRP1 plays a versatile set of roles involving regulation of signaling pathways
+that mediate angiogenesis, axon guidance, cell survival, migration, and cell
+invasion in a wide range of developmental states and tissue types
+\cite{Zachary:2011dr}. It is a single pass transmembrane molecule that has
+multiple extracellular functional domains that interact with VEGF and/or
+semaphorin molecules \cite{Cackowski:2004ja}. NRP1 has been implicated in
+endothelial \cite{Lee:2002dh} and cardiac \cite{Gu:2003vq} development but the
+specific splice variants involved in mediating these processes has not been
+elucidated.  Results shown here indicate that the cardiac splice variant
+contains all functional domains involved in binding VEGF and semaphorin 3a
+(Figure \ref{fig:mesodermal}d-f). This is consistent with studies showing the
+requirement for NRP1 in mediating the activities of VEGF and semaphorins in
+cardiac development \cite{Gu:2003vq}. In contrast, we show that splice variants
+expressed in endothelial populations include truncated versions that either lack
+the semaphorin binding CUB domains and, in the case of HECs, express a secreted
+version of NRP1 (Figure \ref{fig:mesodermal}d-f). This suggests that VEGF and
+not semaphorin 3a is the primary mechanism of signaling through NRP1 in
+endothelial development \cite{Lee:2002dh}. These results provide the first
+evidence to our knowledge that NRP1 splice variants may be key determinants of
+fate determination in mesoderm patterning.
 

### 2016-02-24 supplement.tex
Add Nathan's story. Misc improvements.
Supplement work.

@@
 \documentclass{article}
+\usepackage[left=1in, right=1in]{geometry}
 \usepackage[T1]{fontenc}
@@
 %     isolator.
+%
+%   * Explain the reasoning behind adding 0.1 tpm to expression values to
+%     account for zeros.
 
@@
 the mixed samples C and D and weighted averages of estimates for A and
-B, corresponding to the mixture proportions for C and D. TODO: move this
-table to the supplement.}
+B, corresponding to the mixture proportions for C and D.}
 \end{center}
@@
 
+\section{Notes on measuring agreement between transcript expression estimates}
+
+Benchmarks in this paper rely heavily on quantifying the agreement between
+expression estimates and qPCR, spike-in controls, estimates from other samples,
+or ground truth (in the case of simulations). As discussed in Section
+\ref{results}, our methodology uses proportionality correlation as an
+alternative to more common measures of correlation or aggregate error.
+
+In practice, the distribution of transcript expression values tends to be skewed
+right with a very low median. This can lead to misleading comparisons in two
+ways. First, similarity/dissimilarity measurements like Pearson's correlation or
+root mean square error may be dominated by a small number of very highly
+expressed transcripts. Second, and conversely, measurements like Spearman's rank
+correlation may be dominated by unexpressed, or very low-expression transcripts
+which often make up the large majority of transcripts in a relatively
+well-annotated species like human.
+
+This problem is compounded by the use of different estimators in the methods
+being compared.  When maximum likelihood or, depending on the choice of prior,
+maximum posterior estimates are computed, many expression estimates will be
+zero. On the other hand, posterior mean estimates (as is computed in Isolator,
+BitSeq, and optionally RSEM) are inherently non-zero.
+
+In deciding how to deal with zeros, or the lack of zeros, it is easy to
+artificially advantage one approach or the other. For example, removing zeros,
+disadvantages methods that have no zeros, whereas quantifying error with
+functions like median relative difference (as was used to argue the accuracy of
+Kallisto \cite{Bray:2015uj}), will disadvantage methods that do not produce
+zeros, or otherwise report small numbers.
+\footnote{
+To see this, consider that the absolute relative difference
+between 0 and $\epsilon$ for an arbitrarily large or arbitrarily small but
+positive $\epsilon$ is $\frac{\epsilon - 0}{(\epsilon + 0)/2} = 2$. Thus, MRD
+considers ``absent versus barely present'' to be equivalent to ``absent versus
+highly expressed'', an outcome that is at best, unintuitive.
+
+More generally, the scale-invariance of relative difference means
+
+%% TODO: which benchmark should we use to explore these problems? Maybe the
+% SEQC transcript "consistency" benchmark? Regardless, it seems I didn't save
+% any of this data locally so I have to spin up an instance.
+% Fuck, I've paid about $600 in fees. I need to stop this thing.
+
+$\frac{\alpha |x_1 - x_2|}{\alpha()/2} = 2$
+
+}
+
+This presents two plausible strategies to compensate: truncate small values to
+zero, or add a small constant to every value. Both
+
+\subsection{The effects of truncation}
+
+%This paper argues that posterior mean estimates have some practical benefits.
+%And reporting non-zero estimates in all cases is justifiable from a Bayesian
+%...
+
+\subsection{Effects of varying additive constant on computed proportionality
+            correlation}
+
+As our analysis relies on adding a seemingly arbitrary constant $c$ to
+expression values before computing correlation, a natural question is how
+sensitive our results our to its precise value. In short, we believe it is quite
+insensitive, with results largely consistent when $c$ is varied across several
+orders of magnitude.
+
+%% TODO: plot
+
+Given the insensitivity, we claim that $c = 0.1$ reasonable choice. If we assume
+between $10^5$ to $10^6$ mRNA are present in a mammalian cell
+\cite{Islam:2014is}, then adding $0.1 \text{TPM}$ can be interpreted as
+negligibly inflating expression estimates by 0.01 to 0.1 copies per cell.
+
+\subsection{Select analyses repeated with Spearman's rank correlation}
+
+While we believe proportionality correlation is a more meaningful measure, our
+results remain qualitatively similar using Spearman's rank correlation. In this
+section we reproduce tables from the main paper using this somewhat more
+conventional approach.
+
+\begin{figure*}
+\begin{subtable}{0.5\linewidth}
+\caption{Comparison of estimates to TaqMan qPCR}
+\begin{tabular}[c]{@{}lrrrr@{}}
+\toprule\addlinespace
+Method & A & B & C & D
+\\\addlinespace
+\midrule
+Isolator & \textbf{0.903} & \textbf{0.892} & \textbf{0.878} & \textbf{0.859}
+\\\addlinespace
+eXpress & 0.897 & 0.885 & 0.869 & 0.849
+\\\addlinespace
+Cufflinks & 0.897 & 0.886 & 0.869 & 0.849
+\\\addlinespace
+Salmon  &  0.895 & 0.881 & 0.868 & 0.845
+\\\addlinespace
+RSEM/ML  & 0.895 & 0.880 & 0.867 & 0.844
+\\\addlinespace
+Kallisto &  0.890 & 0.873 & 0.862 & 0.837
+\\\addlinespace
+BitSeq  &  0.886 & 0.874 & 0.862 & 0.842
+\\\addlinespace
+Sailfish &  0.875 & 0.843 & 0.842 & 0.804
+\\\addlinespace
+RSEM/PM  & 0.886 & 0.873 & 0.862 & 0.840
+\\\addlinespace
+\bottomrule
+\addlinespace
+\end{tabular}
+\label{table:taqman:spearman}
+\end{subtable}
+\begin{subtable}{0.5\linewidth}
+\caption{Comparison of estimates to PrimePCR qPCR}
+\begin{tabular}[c]{@{}lrrrr@{}}
+\toprule\addlinespace
+Method & A & B & C & D
+\\\addlinespace
+\midrule
+Isolator  & \textbf{0.872} & \textbf{0.863} & \textbf{0.837} & \textbf{0.844}
+\\\addlinespace
+Cufflinks & 0.864 & 0.855 & 0.826 & 0.835
+\\\addlinespace
+eXpress   & 0.864 & 0.853 & 0.825 & 0.833
+\\\addlinespace
+Salmon    & 0.858 & 0.846 & 0.821 & 0.825
+\\\addlinespace
+RSEM/ML   & 0.856 & 0.845 & 0.819 & 0.823
+\\\addlinespace
+BitSeq    & 0.855 & 0.842 & 0.821 & 0.827
+\\\addlinespace
+Kallisto  & 0.855 & 0.840 & 0.817 & 0.821
+\\\addlinespace
+Sailfish  & 0.840 & 0.811 & 0.797 & 0.793
+\\\addlinespace
+RSEM/PM   & 0.850 & 0.838 & 0.815 & 0.819
+\\\addlinespace
+\bottomrule
+\addlinespace
+\end{tabular}
+\label{table:primepcr:spearman}
+\end{subtable}
+
+\begin{subtable}{0.5\linewidth}
+\caption{Comparison of estimates to ERCC spike-in controls}
+\begin{tabular}[c]{@{}lrrrr@{}}
+\toprule\addlinespace
+Method & A & B & C & D
+\\\addlinespace
+\midrule
+Isolator  & \textbf{0.974} & \textbf{0.973} & \textbf{0.975} & \textbf{0.976}
+\\\addlinespace
+Salmon    & 0.969 & 0.968 & 0.966 & 0.966
+\\\addlinespace
+Kallisto  & 0.964 & 0.964 & 0.961 & 0.965
+\\\addlinespace
+Sailfish  & 0.968 & 0.961 & 0.959 & 0.959
+\\\addlinespace
+Cufflinks & 0.964 & 0.964 & 0.961 & 0.964
+\\\addlinespace
+RSEM/PM   & 0.950 & 0.946 & 0.951 & 0.948
+\\\addlinespace
+RSEM/ML   & 0.945 & 0.950 & 0.948 & 0.948
+\\\addlinespace
+BitSeq    & 0.941 & 0.938 & 0.945 & 0.942
+\\\addlinespace
+eXpress   & 0.937 & 0.944 & 0.941 & 0.942
+\\\addlinespace
+\bottomrule
+\addlinespace
+\end{tabular}
+\label{table:ercc:spearman}
+\end{subtable}
+\begin{subtable}{0.5\linewidth}
+\caption{Consistency of SEQC sample estimates}
+\begin{tabular}[c]{@{}lrr@{}}
+\toprule\addlinespace
+Method & $c$ vs $0.75a + 0.25b$ & $d$ vs $0.25a + 0.75b$
+\\\addlinespace
+\midrule
+Isolator  & \textbf{0.972} & \textbf{0.972}
+\\\addlinespace
+BitSeq    & 0.946 & 0.947
+\\\addlinespace
+RSEM/PM   & 0.936 & 0.935
+\\\addlinespace
+Sailfish  & 0.882 & 0.877
+\\\addlinespace
+RSEM/ML   & 0.841 & 0.838
+\\\addlinespace
+Salmon    & 0.823 & 0.821
+\\\addlinespace
+Kallisto  & 0.845 & 0.842
+\\\addlinespace
+eXpress   & 0.868 & 0.867
+\\\addlinespace
+Cufflinks & 0.865 & 0.864
+\\\addlinespace
+\bottomrule
+\addlinespace
+\end{tabular}
+\label{table:seqctranscripts:spearman}
+\end{subtable}
+
+\caption{
+Tables from Figure 1 recomputed using Spearman's rank correlation.
+}
+\end{figure*}
+
+\begin{table}
+\begin{center}
+\label{table:rlsim:spearman}
+\begin{tabular}[c]{@{}ll@{}}
+\toprule\addlinespace
+Method & Correlation
+\\\addlinespace
+\midrule
+Isolator   & 0.724
+\\\addlinespace
+Kallisto   & 0.727
+\\\addlinespace
+Salmon     & \textbf{0.728}
+\\\addlinespace
+RSEM/ML    & 0.712
+\\\addlinespace
+Cufflinks  & 0.636
+\\\addlinespace
+eXpress    & 0.605
+\\\addlinespace
+Sailfish   & 0.648
+\\\addlinespace
+RSEM/PM    & 0.605
+\\\addlinespace
+BitSeq     & 0.595
+\\\addlinespace
+\bottomrule
+\addlinespace
+\end{tabular}
+\caption{Table 1 recomputed using Spearman's rank correlation.}
+\end{center}
+\end{table}
+
 \end{document}

### 2016-02-27 paper.tex
Supplement work.
Supplemental stuff about propcor.

@@
 
-% TODO: wtf does this break things?
-%\externaldocument{supplement}
+\externaldocument{supplement}
 
@@
 
-\section{Results}
+\section{Results}\label{results}
 
@@
 Zeros are accounted for by converting all measurements to transcripts per
-million (TPM) and adding 0.1.
-% TODO: add supplementary info about why 0.1 is a reasonable value.
+million (TPM) and adding 0.1. Results were found to be fairly insensitive to the
+precise additive constant used here (Supplementary Section
+\ref{additive-constant}). Though we believe proportionality correlation is a
+superior approach, much of this analysis was also conducted using Spearman's
+rank correlation in Supplementary Section \ref{spearman-reanalysis}, and largely
+agrees with the results shown here.
 
@@
 Kallisto.
-% TODO: see supplementary section XXX
+% TODO: see supplementary section XXX (how reads were aligned)
 
@@
 \midrule
-\textbf{Isolator} & \textbf{0.919}
+Isolator & \textbf{0.919}
 \\\addlinespace

### 2016-02-27 supplement.tex
Supplemental stuff about propcor.

@@
 
-Existing statisticaly models (see {[}@Pachter:2011wm{]} for a review),
+Existing statisticaly models (see \cite{Pachter:2011wm} for a review),
 assume fragments are sampled uniformly at random from a transcript. This
@@
 \begin{figure}[H]
+\centering
 \includegraphics[width=0.2\textwidth]{analysis/cardio/colorkey.pdf} \\
-\includegraphics[width=\textwidth]{analysis/cardio/ACOT9-comparison.pdf}
-\includegraphics[width=\textwidth]{analysis/cardio/IMMT-comparison.pdf}
-\includegraphics[width=\textwidth]{analysis/cardio/LTBP4-comparison.pdf}
+\includegraphics[width=0.9\textwidth]{analysis/cardio/ACOT9-comparison.pdf}
+\includegraphics[width=0.9\textwidth]{analysis/cardio/IMMT-comparison.pdf}
 \end{figure}
@@
 \ContinuedFloat
-\includegraphics[width=\textwidth]{analysis/cardio/OBSL1-comparison.pdf}
-\includegraphics[width=\textwidth]{analysis/cardio/OSBPL1A-comparison.pdf}
-\includegraphics[width=\textwidth]{analysis/cardio/PALM-comparison.pdf}
-\includegraphics[width=\textwidth]{analysis/cardio/TNNT2-comparison.pdf}
+\centering
+\includegraphics[width=0.9\textwidth]{analysis/cardio/LTBP4-comparison.pdf}
+\includegraphics[width=0.9\textwidth]{analysis/cardio/OBSL1-comparison.pdf}
+\includegraphics[width=0.9\textwidth]{analysis/cardio/OSBPL1A-comparison.pdf}
+\end{figure}
+\begin{figure}[H]
+\ContinuedFloat
+\centering
+\includegraphics[width=0.9\textwidth]{analysis/cardio/PALM-comparison.pdf}
+\includegraphics[width=0.9\textwidth]{analysis/cardio/TNNT2-comparison.pdf}
 \caption{Point estimates of fold change of exon expression from qPCR and and
@@
 or ground truth (in the case of simulations). As discussed in Section
-\ref{results}, our methodology uses proportionality correlation as an
-alternative to more common measures of correlation or aggregate error.
+\ref{results}, our methodology uses proportionality correlation
+\cite{Lovell:2015il} as an alternative to more common measures of correlation or
+aggregate error.
 
@@
 
-This problem is compounded by the use of different estimators in the methods
-being compared.  When maximum likelihood or, depending on the choice of prior,
-maximum posterior estimates are computed, many expression estimates will be
-zero. On the other hand, posterior mean estimates (as is computed in Isolator,
-BitSeq, and optionally RSEM) are inherently non-zero.
+This problem is not adressed by using median relative difference (MRD), as was used to
+argue the accuracy of Kallisto \cite{Bray:2015uj}. Though it may seem robust due
+to the use of a median, it does not account for the compositional nature of
+RNA-Seq abundance estimates. A large error in one highly expressed transcript
+can inflate the relative error in all other transcripts, thus driving up the
+median. Furthermore, relative error can be unintuitive in the presense of
+zero estimates. To see this, consider that the relative difference
+between 0 and $\epsilon$ for an arbitrary positive $\epsilon$ is $\frac{\epsilon
+- 0}{(\epsilon + 0)/2} = 2$. Thus, MRD unintuitively treats ``absent versus
+barely present'' to be equivalent to ``absent versus highly expressed''.
 
-In deciding how to deal with zeros, or the lack of zeros, it is easy to
-artificially advantage one approach or the other. For example, removing zeros,
-disadvantages methods that have no zeros, whereas quantifying error with
-functions like median relative difference (as was used to argue the accuracy of
-Kallisto \cite{Bray:2015uj}), will disadvantage methods that do not produce
-zeros, or otherwise report small numbers.
-\footnote{
-To see this, consider that the absolute relative difference
-between 0 and $\epsilon$ for an arbitrarily large or arbitrarily small but
-positive $\epsilon$ is $\frac{\epsilon - 0}{(\epsilon + 0)/2} = 2$. Thus, MRD
-considers ``absent versus barely present'' to be equivalent to ``absent versus
-highly expressed'', an outcome that is at best, unintuitive.
-
-More generally, the scale-invariance of relative difference means
-
-%% TODO: which benchmark should we use to explore these problems? Maybe the
-% SEQC transcript "consistency" benchmark? Regardless, it seems I didn't save
-% any of this data locally so I have to spin up an instance.
-% Fuck, I've paid about $600 in fees. I need to stop this thing.
-
-$\frac{\alpha |x_1 - x_2|}{\alpha()/2} = 2$
-
-}
-
-This presents two plausible strategies to compensate: truncate small values to
-zero, or add a small constant to every value. Both
-
-\subsection{The effects of truncation}
-
-%This paper argues that posterior mean estimates have some practical benefits.
-%And reporting non-zero estimates in all cases is justifiable from a Bayesian
-%...
+For these reasons, we believe proportionality correlation is a more appropriate
+measure when evaluating relative expression estimates.  The prevailing practical
+issue with proportionality correlation is that it is defined only over positive
+numbers. Methods that use maximum likelihood (or, depending on the choice of
+priors, posterior maximum) estimates do report zeros.  We account for this by
+adding a small constant ($c = 0.1 \text{TPM}$) to all expression values before
+computing proportionality correlation.
 
 \subsection{Effects of varying additive constant on computed proportionality
-            correlation}
+            correlation}\label{additive-constant}
 
@@
 insensitive, with results largely consistent when $c$ is varied across several
-orders of magnitude.
+orders of magnitude (Supplementary Figure \ref{fig:propcor-constant}).
 
-%% TODO: plot
+\begin{figure*}
+    \centering
+    \begin{subfigure}[b]{0.8\textwidth}
+    \centering
+    \includegraphics[width=0.8\textwidth]{pc-constant-seqc-primepcr.pdf}
+    \caption{}
+    \end{subfigure}
+
+    \begin{subfigure}[b]{0.8\textwidth}
+    \centering
+    \includegraphics[width=0.8\textwidth]{pc-constant-seqc-transcript-expression-consistency.pdf}
+    \caption{}
+    \end{subfigure}
+
+    \caption{To account for zeros and near-zero values, a small constant $c$ is
+        added to expression values before computing proportionality correlation.
+        Throughout the paper we use $c = 0.1 \text{TPM}$. Here we varied the
+        value of $c$ across several orders of magnitude and recomputed
+        proportionality correlation in two of the benchmarks used in the paper:
+        (a) comparison to PrimePCR qPCR and (b) consistency of transcript
+        expression in SEQC data. Though the computed correlation changes
+        depending on the value of $c$, methods largely maintain the same order.
+    }
+    \label{fig:propcor-constant}
+\end{figure*}
 
@@
 
-\subsection{Select analyses repeated with Spearman's rank correlation}
+\subsection{Select analyses repeated with Spearman's rank correlation}\label{spearman-reanalysis}
 
@@
 \caption{
-Tables from Figure 1 recomputed using Spearman's rank correlation.
+Results from Figure 1 from the main paper recomputed using Spearman's rank correlation.
 }
@@
 \end{tabular}
-\caption{Table 1 recomputed using Spearman's rank correlation.}
+\caption{Table 1 from the main paper recomputed using Spearman's rank correlation.}
 \end{center}
@@
 
+\bibliographystyle{plainnat}
+\bibliography{references,extra-references}
+
 \end{document}

### 2016-05-31 paper.tex
Minor fixes. Author affiliations.

@@
 \author[1]{Daniel C. Jones}
-\author[2]{Kavitha T. Kuppusamy}
+\author[2,6]{Kavitha T. Kuppusamy}
 \author[3]{Nathan J. Palpant}
@@
 \author[4]{Michael G. Katze}
-\author[5]{Charles E. Murry}
-\author[6]{Hannele Ruohola-Baker}
+\author[2,5]{Charles E. Murry}
+\author[2,6]{Hannele Ruohola-Baker}
 \author[1,7,8]{Walter L. Ruzzo}
@@
 \affil[1]{Department of Computer Science and Engineering, University of Washington}
-\affil[2]{XXX}
-\affil[3]{XXX}
+\affil[2]{Institute for Stem Cell and Regenerative Medicine}
+%\affil[3]{Institute for Molecular Bioscience, The University of Queensland. Brisbane. Australia}
+\affil[3]{Institute for Molecular Bioscience, The University of Queensland}
 \affil[4]{Department of Microbiology, University of Washington}
@@
 splicing rarely concuring \cite{Liu:2014dn}.
+
+Regardless, splicing has been repeatedly shown to be a essential regulatory
+mechanism, and particularly important in development and differentiation
+\cite{Kalsotra:2011hv}. We examine alternative splicing in heart development in
+two contexts in Sections \ref{finding-monotonic-differential-splicing} and
+\ref{novel-mechanisms-of-cell-fate}.
 
@@
 which notably excludes the alignment-free methods Sailfish, Salmon, and
-Kallisto.
-% TODO: see supplementary section XXX (how reads were aligned)
+Kallisto (Supplementary Section \ref{analysis}).
 
@@
 Beijing Genome Institute (BGI), Cornell University (CNL), Mayo Clinic (MAY), and
-Novartis (NVS). \textbf{b} The absolute change in correlation induced by
-enabling bias correction for methods that support it. For clarity this plot
-excludes points for BitSeq estimates of "MAY 2", as bias correction has an
-extremely detrimental effect on these. Mean improvement in correlation was 0.008
-with Salmon, 0.007 with Cufflinks, 0.006 with Isolator, 0.003 with Kallisto,
-0.002 with eXpress, and -0.164 with BitSeq.
+Novartis (NVS). Median proportionality correlation is listed below each heatmap.
+\textbf{b} The absolute change in correlation induced by enabling bias
+correction for methods that support it. For clarity this plot excludes points
+for BitSeq estimates of "MAY 2", as bias correction has an extremely detrimental
+effect on these. Mean improvement in correlation was 0.008 with Salmon, 0.007
+with Cufflinks, 0.006 with Isolator, 0.003 with Kallisto, 0.002 with eXpress,
+and -0.164 with BitSeq.
 }
@@
 instruments, and compared pairwise agreement between estimates from these
-samples (Figure \ref{fig:batcheffects}a).
+samples (Figure \ref{fig:batcheffects}a). In the interest of a more fair
+comparison, Isolator was run on each sample in isolation without sharing
+any information.
 
@@
 \begin{table}
-\label{table:rlsim}
 \begin{tabular}[c]{@{}ll@{}}
@@
 produced by each method using simulated RNA-Seq reads.}
+\label{table:rlsim}
 \end{table}
@@
 shows extremely high correlation with small numbers of reads, but
-contrary to all other methods, correlation rapidly decreases with deeper
-sequencing after 1 million reads. When posterior mean estimates are
+contrary to all other methods, correlation rapidly decreases with
+sequencing deeper than 1 million reads. When posterior mean estimates are
 generated from a model using a uninformative prior, as is the case with
@@
 sequencing depth, though the correlation remains very low. Even using
-all 11 million reads, RSEM/PM had a correlation of only 0.567.
+all 11 million reads, RSEM/PM had a correlation of only 0.567 (data not shown).
 
@@
 
-Splicing has been known to play an important role in devoloping hearts. Specific
+Splicing has been known to play an important role in developing hearts. Specific
 splicing changes during maturation have been shown to be conserved across
@@
 probability of ``monotonic splicing``, which we defined as a splicing change
-that that occurs is consistently observed as occuring in the same direction
+that that is consistently observed to occur in the same direction
 between immature (H7 cells at day 20 and at day 30 infected with an empty
@@
 OBSL1, OSBPL1A, PALM, and TNNT2) to validate with qPCR (Supplementary Section
-\ref{cardioqpcr}), by measuring the expression of specific exons predicted a
+\ref{cardioqpcr}), by measuring the expression of specific exons predicted to
 change in splicing. Five of these candidates (ACOT9, LTBP4, OBSL1, OSBPL1A, and
@@
 \subsection{Case study: novel mechanisms of cell fate specification identified
-by gene splice variants}
+by gene splice variants}\label{novel-mechanisms-of-cell-fate}
 
@@
 Analysis of gene ontology showing enriched categories found in 84 differentially
-spliced genes. (d-e) Heat map analysis 
+spliced genes.
 \textbf{d}
@@
 the same alignments for every method (excluding the alignment-free methods
-Salmon, Sailfish, and Kallisto), and compare wider variety methods. In this
+Salmon, Sailfish, and Kallisto), and compare a wider variety methods. In this
 broader context it becomes clear that this separation is a consequence of the
@@
 accurate, frequently demonstrate lower consistency than methods that use MCMC
-sampling and report posterior means or medians. This should be unsurprising.
-Because of the ambiguity of sequencing data, the likelihood functions for
-relative expression are often near plateaus. Maximum likelihood estimates are
-thus are precarious compared to posterior mean estimates. Neither estimate is
-inherently more correct, and robust analysis should make use of the full
-posterior distribution, but we argue that posterior mean or median estimates
-are more useful in practice because of this property.
+sampling and report posterior means or medians. This is underappreciated but
+should be unsurprising.  Because of the ambiguity of sequencing data, the
+likelihood functions for relative expression are often near plateaus. Maximum
+likelihood estimates are thus precarious compared to posterior mean
+estimates. Neither estimate is inherently more correct, and robust analysis
+should make use of the full posterior distribution, but we argue that posterior
+mean or median estimates are more useful in practice because of their greater
+stability.
 
-Most methods that report posterior mean estimates show seemingly poor accuracy.
-Much of the cause of the seemingly reduced accuracy of the MCMC based models is
-driven by low-expression transcripts. A transcript with no reads has a maximum
-likelihood expression of zero, but likelihood is positive over a range of
-non-zero values, so the posterior mean must necessarily also be positive. With
-a flat, or ``non-informative'' prior expression values for these low-expression
-transcripts are often quite high and highly determined by transcript length.
-Isolator avoids this problem by using a model with informative priors: namely
-encoding assumptions that most transcripts have low expression and that
-expression and splicing tend to be similar between samples. In this way, it is
-able to avoid what seems like a tradeoff and produce estimates that are both
-consistent and
-accurate.
+Excluding Isolator, methods that report posterior mean estimates show seemingly
+poor accuracy. This is largely driven by low-expression transcripts. A
+transcript with no reads has a maximum likelihood expression of zero, but its
+likelihood is positive over a range of non-zero values, so the posterior mean
+must necessarily also be positive. With a flat, or ``non-informative,'' prior,
+expression values for these low-expression transcripts are often quite high and
+highly determined by transcript length.  Isolator avoids this problem by using a
+model with informative priors: namely encoding assumptions that most transcripts
+have low expression and that expression and splicing tend to be similar between
+samples. In this way, it is able to avoid what seems like a tradeoff and
+produces estimates that are both consistent and accurate.
 
 Fully Bayesian methods like the one presented here possess a degree of
-subjectivity that sometimes give researchers pause. Although we have
-found the results from Isolator to be insensitive to precise values
-given to hyperparameters, they are chosen in advance and with a degree
-of arbitrariness. What is often ignored is that the alternative, methods
-based on unregularized maximum likelihood estimates, in practice often
-necessitate a more insidious form of subjectivity: contriving cutoffs to
-filter out unreliable point estimates of low expression genes or
-transcripts. While the priors in Bayesian methods are purely explicit,
-the means by which data was filtered during an analysis is often a form
-of off-the-books subjectivity--critical to the results, yet unmentioned
-or only alluded to in manuscripts.
+subjectivity that sometimes give researchers pause. Although we have found the
+results from Isolator to be insensitive to precise values given to
+hyperparameters, they are chosen in advance and with a degree of arbitrariness.
+What is often ignored is that the alternative, methods based on unregularized
+maximum likelihood estimates, in practice often necessitate a more insidious
+form of subjectivity: contriving cutoffs to filter out unreliable point
+estimates of low expression genes or transcripts. While the priors in Bayesian
+methods are purely explicit, ad hoc filtering of unregularized estimates can
+often be a form of off-the-books subjectivity--critical to the results, yet
+unmentioned or only alluded to in manuscripts.
+
+In this paper we have described a method that effectively shares information
+between samples in RNA-Seq experiments while rigorously quantifying uncertainty,
+leading to more accurate results across a wide range of benchmarks. We believe
+this work indicates a natural direction towards more reliable and powerful
+studies of transcriptomics, which future work can follow by scaling this
+methodology to larger numbers of samples, and accounting for more forms of
+uncertainty and technical effects.
 
@@
 
+Input to Isolator consists of gene annotations in GTF/GFF format, one
+or more sorted BAM \cite{Li:2009ka} file, and optionally a reference sequence,
+which if provided is used in bias correction.
+
 To model RNA-Seq experiments we use a hierarchical model consisting of
-three levels: samples or replicates, conditions, and the experiment. Any
+three levels: samples or replicates, conditions, and the experiment, with any
 number of conditions, and any number of replicates within those
@@
 
+Parameters are updated in sequence using a Gibbs sampling approach. The output
+generated by Isolator consists of each sample of every parameter serialized in
+an HDF5-based format, along with basic diagnostic information.  Point estimates,
+summary statistics, and posterior probabilities can then be calculated from this
+output using a secondary command (\texttt{isolator summarize}).
+
 \subsection{Efficient Sampling}\label{efficient-sampling}
@@
 (isoform abundances) must be inferred. The standard approach to this
-problem is ether expectation maximization (EM) if the maximum likelihood
+problem is either expectation maximization (EM) if the maximum likelihood
 solution is desired, or Gibbs sampling if, as we are, estimating the
 posterior distribution. We have taken a different approach and instead
-rely on slice sampling.
+rely on slice sampling \cite{Neal:2000uc}.
 
@@
 autocorrelation in a Gibbs sampler by drawing more samples in less time.
-It is not obvious then that one approach is inherently more efficient
+It is not obvious, then, that one approach is inherently more efficient
 than the other.
 
-Because the likelihood function in this problem is simple, and we sample
-from the posterior probability directly, without introducing latent
-variables, we are afforded some optimization opportunities not available
-in a Gibbs sampler. Specifically, we implement the likelihood function
-using SIMD (single instruction, multiple data) instructions, which are
-available nearly all CPUs from the last decade. This, in combination
-with some numerical approximations, allow us to compute the likelihood
-function over an order of magnitude faster on conventional hardware,
+Because the likelihood function in this problem is simple, and we sample from
+the posterior probability directly, without introducing latent variables, we are
+afforded some optimization opportunities not available in a Gibbs sampler.
+Specifically, we implement the likelihood function using SIMD (single
+instruction, multiple data) instructions, which are available nearly all CPUs
+from the last decade. This, in combination with some numerical approximations,
+allow us to compute the likelihood function over an order of magnitude faster,
 than a more straightforward implementation. Run time for Isolator is generally
@@
 
-Perhaps the most prominent source of nonuniformity is sequence specific
-bias present at fragment ends. A probable source of this bias is cDNA
-synthesis by random priming, which is part of popular RNA-Seq protocols
-such as Illumina's TruSeq kit. We have previously published a model that
-successfully accounts for effect by training a sparse graphical model
-\cite{Jones:2012hg}.
+Perhaps the most prominent source of nonuniformity is sequence specific bias
+present at fragment ends. A probable contributor of this bias is cDNA synthesis
+by random priming, which is part of popular RNA-Seq protocols such as Illumina's
+TruSeq kit. We have previously published a model that successfully accounts for
+such effects by training a sparse graphical model \cite{Jones:2012hg}.
 
-Beyond this random priming bias, there is residual fragment GC-bias,
+Beyond this sequence context bias, there is residual fragment GC-bias,
 which may be an artifact of PCR amplification. We observe that fragments
-with very high or very low GC-content are less likely to be observed
-than expected, even after accounting for priming bias. We model this by
+with very high or very low GC-content are less likely to be sequenced
+than expected, even after accounting for sequence bias. We model this by
 binning fragments by GC content, computing expected and observed
@@
 
-Selection of polyadenylated transcripts is a common step mRNA-Seq used
-to avoid sequencing introns, partially degraded transcripts, and
-ribosomal RNA. This can cause enrichment of reads at the 3' end of
-transcripts if only partial transcripts are captured. To model the
-effect we fit a one parameter model in which a transcript is truncated
-at any position with probability $p$. The probability of observing a
-fragment at a position $k$ of length $\ell$, counting from the 3' end,
-is then proportional to the probability that the transcript was
-\emph{not} truncated before $k + \ell$, which is $(1-p)^{k + \ell}$.
-When fit, $p$ is typically quite small, on the order of $10^{-5}$, in
-which case this correction has little effect on transcripts shorter that
-several kilobases.
+Selection of polyadenylated transcripts is a common step mRNA-Seq, used to avoid
+sequencing introns, partially degraded transcripts, and ribosomal RNA. This can
+cause enrichment of reads at the 3' end of transcripts if only partial
+transcripts are captured. To model the effect we fit a one parameter model in
+which a transcript is truncated at any position with probability $p$. The
+probability of observing a fragment ending at a position $k$, counting from the
+3' end, is then proportional to the probability that the transcript was
+\emph{not} truncated before $k$, which is $(1-p)^{k}$.  When fit, $p$ is
+typically quite small, on the order of $10^{-5}$, in which case this correction
+has little effect on transcripts shorter that several kilobases.
 
@@
 fragments at either end of the transcript. We compute this enrichment exactly
-for various transcript length, and use interpolation to approximate th effect.
+for various transcript length, and use interpolation to approximate the effect.
 

### 2016-06-01 paper.tex
NextSeq not HiSeq was used in rhemac data.

@@
 paired-end reads on a MiSeq sequencer and again with 100nt paired-end
-reads using a HiSeq 2000, yielding approximately 4.5 million and 11
+reads using a NextSeq 500, yielding approximately 4.5 million and 11
 million reads, respectively. To elucidate the effect of deeper
 sequencing, we sampled without replacement smaller subsets of the 11
-million HiSeq 2000 reads and measured correlation between estimates made using
+million NextSeq 500 reads and measured correlation between estimates made using
 these subsamples with estimates from the MiSeq data (Figure \ref{fig:rhemac}).
@@
 \caption{Proportionality correlation between estimates from 4.5 million 300nt
-MiSeq reads and progressively larger numbers of HiSeq 2000 reads. For clarity,
+MiSeq reads and progressively larger numbers of NextSeq 500 reads. For clarity,
 correlations for RSEM/PM are excluded as they were much lower than the other

### 2016-09-23 paper.tex
Try to solidify discussions of consistency/stability.

@@
     level. When confronted with this uncertainty, commonly used methods produce
-    estimates that are unstable---small perturbations in the data often produce
-    dramatically different results, confounding downstream analysis. We
-    introduce a new method, Isolator, which analyzes all samples in an
-    experiment in unison using a simple Bayesian hierarchical model. Combined
+    estimates that are often high-variance---small perturbations in the data
+    often produce dramatically different results, confounding downstream
+    analysis. We introduce a new method, Isolator, which analyzes all samples in
+    an experiment in unison using a simple Bayesian hierarchical model. Combined
     with aggressive bias correction, it produces estimates that are
     simultaneously accurate and stable. In a comprehensive comparison of
-    accuracy and stability, we show that this property is unique to Isolator. We
+    accuracy and variance, we show that this property is unique to Isolator. We
     further demonstrate that the approach of modeling an entire experiment
@@
 
-We built on this work, implementing a new method called Isolator. Though
-it retains the same underlying model of RNA-Seq, we approach the problem
-from the perspective that isoform-level estimates possess a degree of
-uncertainty that renders maximum likelihood point-estimates alone a
-fundamentally inadequate solution to the problem. As the purpose of gene
-expression studies is to compare expression between multiple samples,
-our goal, beyond maximizing a narrow definition of accuracy, is to
-provide a tool that effectively accounts for this uncertainty in a
-coherent and reliable manner.
+We built on this work, implementing a new method called Isolator. Though it
+retains the same underlying model of RNA-Seq, we approach the problem from the
+perspective that isoform-level estimates possess a degree of uncertainty that
+renders unregularized maximum likelihood point-estimates alone a fundamentally
+inadequate solution to the problem. As the purpose of gene expression studies is
+to compare expression between multiple samples, our goal, beyond maximizing a
+narrow definition of accuracy, is to provide a tool that effectively accounts
+for this uncertainty in a coherent and reliable manner.
 
@@
 estimation exists, we adopt a multifaceted approach, focusing on measuring
-the stability or consistency of estimates, along with accuracy.
+the variance of estimates, along with overall accuracy.
 
@@
 to other measures of correlation, proportionality correlation varies from -1 to
-1, with 1 indicating perfectly proportionality, and -1 perfect reciprocality.
+1, with 1 indicating perfect proportionality, and -1 perfect reciprocality.
 Zeros are accounted for by converting all measurements to transcripts per
-million (TPM) and adding 0.1. Results were found to be fairly insensitive to the
+million (TPM) and adding 0.1. (Results were found to be fairly insensitive to the
 precise additive constant used here (Supplementary Section
-\ref{additive-constant}). Though we believe proportionality correlation is a
+\ref{additive-constant})). Though we believe proportionality correlation is a
 superior approach, much of this analysis was also conducted using Spearman's
@@
 exhaustive account of such methods, but does represent a wide variety of popular
-approaches. When possible, the same read alignments were used for each method,
-which notably excludes the alignment-free methods Sailfish, Salmon, and
-Kallisto (Supplementary Section \ref{analysis}).
+approaches. The same read alignments were used for all methods other Sailfish,
+Salmon, and Kallisto, which are ``alignment-free'', using built-in
+approximations to full alignment (Supplementary Section \ref{analysis}).
 
@@
 \caption{
-Benchmarks measuring the accuracy and consistency of estimates made from
+Benchmarks measuring the accuracy and coherence of estimates made from
 Sequencing Quality Control (SEQC) dataset using a variety of methods.
@@
 
-\subsection{Estimation consistency}\label{estimation-consistency}
+\subsection{Estimate coherence}\label{estimation-coherence}
 
@@
 $B$ in specific proportions. Sample $C$ is composed of 75\% $A$ and 25\% $B$,
-while sample $D$ is 25\% $A$ and 75\% $B$. This design allows us to directly
-measure expression consistency. If we estimate transcript abundance for each
-sample, producing estimates $a, b, c, d$, then for these estimates to be in
-agreement we should expect $c$ to be approximated by $0.25a + 0.75b$, and
-similarly $d$ by $0.75a + 0.25b$. To capture the notion of consistency, we
-measured the closeness of this approximation using proportionality correlation
-(Figure \ref{table:seqctranscripts}).
+while sample $D$ is 25\% $A$ and 75\% $B$. If we estimate transcript abundance
+for each sample, producing estimates $a, b, c, d$, then for these estimates to
+be in agreement (``coherent'', in our terminology) we should expect $c \appprox
+0.25a + 0.75b$, and similarly $d \approx 0.75a + 0.25b$. In other words, mixing
+RNA then estimating transcript abundance should be approximated by estimating
+transcript abundance then mixing estimates.
+
+To capture the notion of coherence, we measured the closeness of this
+approximation using proportionality correlation (Figure
+\ref{table:seqctranscripts}).
 
@@
 
-We further examined the question of consistency by comparing the same
-samples sequenced on different flow cells and at different sites, again
-using data from the SEQC. We used 10 samples from the larger SEQC
-dataset, each consisting of a single lane from two separate
-flowcells sequenced at five different sites, all on Illumina HiSeq 2000
-instruments, and compared pairwise agreement between estimates from these
-samples (Figure \ref{fig:batcheffects}a). In the interest of a more fair
-comparison, Isolator was run on each sample in isolation without sharing
-any information.
+We further examined the question of estimate variance by comparing the same
+samples sequenced on different flow cells and at different sites, again using
+data from the SEQC. We used 10 samples from the larger SEQC dataset, each
+consisting of a single lane from two separate flowcells sequenced at five
+different sites, all on Illumina HiSeq 2000 instruments, and compared pairwise
+agreement between estimates from these samples (Figure \ref{fig:batcheffects}a).
+In the interest of a more fair comparison, Isolator was run on each sample in
+isolation without sharing any information.
 
@@
 alternative methods using a wide variety of benchmarks, with a particular focus
-on estimation consistency or stability. Data generated by the SEQC is ideally
-suited to examine the question consistency, and indeed our data agrees with
-their results. The SEQC compared Cufflinks and BitSeq and found BitSeq to
-produce more consistent estimates. Here we control for alignment method by using
-the same alignments for every method (excluding the alignment-free methods
-Salmon, Sailfish, and Kallisto), and compare a wider variety methods. In this
-broader context it becomes clear that this separation is a consequence of the
-choice in estimators. Unmoderated maximum likelihood methods, though often
-accurate, frequently demonstrate lower consistency than methods that use MCMC
-sampling and report posterior means or medians. This is underappreciated but
-should be unsurprising.  Because of the ambiguity of sequencing data, the
-likelihood functions for relative expression are often near plateaus. Maximum
-likelihood estimates are thus precarious compared to posterior mean
-estimates. Neither estimate is inherently more correct, and robust analysis
-should make use of the full posterior distribution, but we argue that posterior
-mean or median estimates are more useful in practice because of their greater
-stability.
+on estimation variance and coherence. Data generated by the SEQC is ideally
+suited to examine this question, and indeed our analysis agrees with their
+results. The SEQC compared Cufflinks and BitSeq and found BitSeq to produce more
+consistent, less variable estimates. We took a closer look at this phenomenon,
+controlling for alignment method by using the same set of alignments for every
+method (excluding the alignment-free methods Salmon, Sailfish, and Kallisto),
+and comparing a wider variety of methods. In this broader context it becomes
+clear that this separation is a consequence of the choice in estimators.
+Unmoderated maximum likelihood methods, though often accurate (i.e. low bias),
+frequently demonstrate less consistency (i.e. high variance) than methods that
+use MCMC sampling and report posterior means or medians. This is
+underappreciated but should be unsurprising.  Because of the ambiguity of
+sequencing data, the likelihood functions for relative expression are often
+nearly plateaus. Maximum likelihood estimates are thus precarious compared to
+posterior mean estimates. Neither estimate is inherently more correct, and
+robust analysis should make use of the full posterior distribution, but we argue
+that posterior mean or median estimates are more useful in practice because they
+tend to have lower variance.
 
@@
 samples. In this way, it is able to avoid what seems like a tradeoff and
-produces estimates that are both consistent and accurate.
+produces estimates that are both relatively low bias and low variance.
 
@@
 studies of transcriptomics, which future work can follow by scaling this
-methodology to larger numbers of samples, and accounting for more forms of
-uncertainty and technical effects.
+methodology to larger numbers of samples, allowing for more flexible models, and
+accounting for more forms of uncertainty and technical effects.
 

### 2016-11-30 paper.tex
Nix Katze
tweak some terminology

@@
 
-\title{Isolator: accurate and consistent analysis of isoform-level expression in RNA-Seq experiments}
+\title{Isolator: accurate and stable analysis of isoform-level expression in RNA-Seq experiments}
 
@@
 \author[4]{Xinxia Peng}
-\author[4]{Michael G. Katze}
+%\author[4]{Michael G. Katze}
 \author[2,5]{Charles E. Murry}
@@
     with aggressive bias correction, it produces estimates that are
-    simultaneously accurate and stable. In a comprehensive comparison of
-    accuracy and variance, we show that this property is unique to Isolator. We
-    further demonstrate that the approach of modeling an entire experiment
-    enables new analyses, which we demonstrate by examining splicing
-    monotonicity across several time points in the development of human
-    cardiomyocyte cells.
+    simultaneously accurate and show high agreement between samples. In a
+    comprehensive comparison of accuracy and variance, we show that this
+    property is unique to Isolator. We further demonstrate that the approach of
+    modeling an entire experiment enables new analyses, which we demonstrate by
+    examining splicing monotonicity across several time points in the
+    development of human cardiomyocyte cells.
 \end{abstract}
@@
 for each sample, producing estimates $a, b, c, d$, then for these estimates to
-be in agreement (``coherent'', in our terminology) we should expect $c \appprox
+be in agreement (``coherent'', in our terminology) we should expect $c \approx
 0.25a + 0.75b$, and similarly $d \approx 0.75a + 0.25b$. In other words, mixing
@@
 approximately 5\%. This stands in contrast to comparisons to qPCR which suggest
-reduced accuracy of posterior mean estimates, demonstrating that consistency and
+reduced accuracy of posterior mean estimates, demonstrating that coherence and
 accuracy, while related, are two separate axis of comparison.
